<span style="display: none;">
$$
\newcommand{\Lagr}{\mathcal{L}}
\newcommand{\bP}{\mathbb{P}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\bQ}{\mathbb{Q}}
\newcommand{\bE}{\mathbb{E}}
\Lagr \bP \bR \bQ \bE 
$$ 
<span style="display: none;">

# Financial Derivatives {#FD}

**Financial Derivatives** are securities that derive their value from
cash markets. In Ingersoll's definition, a financial contract is a
derivative security.

**Cash and Carry Markets**: Gold, silver, T-bonds. The idea is to borrow
cash, buy asset, and hold until expiration.

**Forward Contract**: obligation to buy/sell an asset at a specified
price at a future date. if cash price is higher at expiration then
forward, and long position make a profit.

```{r, echo = FALSE}
ForwardContracts <- cbind(c( "Mark to Market", "Settlement", "Margin System", "Trading", "Types", "Flexibility", "Liquidity", "Counterparty Risk"),
                          c( "Yes", "Daily", "Yes", "Exchanges", "Standardized" , "Less", "More", "Less"),
                          c("No", "Upon Maturity", "No", "OTC", "Custom", "More", "Less", "More")
                          )

kableExtra::kbl(ForwardContracts, col.names = c("Characteristic", "Futures", "Forwards"), format = "html") %>% kable_minimal("hover")


```

**Repurchase Agreement (REPO)**: A transaction in which one party sells
a security in return for cash, with an agreement to repurchase the
security at an agreed upon price at a future date.

[**Options**](#Options):

-   European: The right to buy $S_t$ at strike k. This right may be
    excercised at time T. Where $S_T <K$ is out the money and $S_T > K$
    is in the money. $$C_T = \max(S_T -K)$$

**Swaps**: Exchange of one set of cashflows for another. Swaps are
simultaneous and can involve curruncies, interst rates, etc. Swaps can
be priced through forwards. Cancelable swaps allow the right to cancel
swap contracts.

**Arbitrage** : Taking simultaneous positions so that it guarantees a
riskless profit higher than the risk-free rate. Let $d_{ij}$ be the
number of units of account paid by one unit of security i in state j.
Rows represent asset, columns represent state (interest, conflict,
recession, etc...)


::: columns
::: {.column width="20%"}
:::

::: {.column width="30%"}
$$
\text{Asset Values} = S_T = 
\begin{bmatrix} 
S_1(t) \\
S_2(t) \\
\vdots \\
S_n(t)
\end{bmatrix}
$$
:::

::: {.column width="30%"}
$$
D_t = \begin{bmatrix}
d_{11} & d_{12} & \cdots & d_{1k} \\
d_{21} & d_{22} & & d_{2k} \\
\vdots & & \ddots & \vdots\\
d_{i1} & d_{i2} & \cdots & d_{ik}
\end{bmatrix}
$$
:::

::: {.column width="20%"}
:::
:::

$$
\text{example of a 2 state portfolio with cash, security, and call}\\ \begin{bmatrix}
(1+r\Delta) & (1+r\Delta) \\
S_1(t + \Delta) & S_2(t + \Delta) \\
C_1(t + \Delta) & C_2(t + \Delta) 
\end{bmatrix}
$$

No arbitrage possibility exists iff:

$$
\begin{bmatrix}
1 \\ S(t) \\ C(t)
\end{bmatrix} 
=
\begin{bmatrix}
(1+r\Delta) & (1+r\Delta) \\
S_1(t + \Delta) & S_2(t + \Delta) \\
C_1(t + \Delta) & C_2(t + \Delta) 
\end{bmatrix} 
\begin{bmatrix}
\psi_1 \\ \psi_2
\end{bmatrix} 
$$

$\psi_n$ represents the risk neutral probability discounted at the risk
free rate and:

$$ \sum_i^n \psi_i = v = \frac{1}{1+r} \rightarrow \sum_i^n \psi_i \cdot (1+r) = \sum_i^n  \bQ_i  = 1 $$

|       **Binomial Tree Pricing**: A method of valuing options. Steps to determine option value is to first determine the value at expiration, then work backwards each step until the current price is obtained.

$$
C_T = \frac{1}{1+r}[Q_{up} C_{t+\Delta}^{up} + Q_{down} C_{t+\Delta}^{down}]
$$

```{tikz, BinomialTree, fig.ext = 'png', cache=TRUE, echo = FALSE, fig.dim = c(5, 4), fig.align="center"}
  \begin{tikzpicture}[>=stealth,sloped]
    \matrix (tree) [%
      matrix of nodes,
      minimum size=1cm,
      column sep=3.5cm,
      row sep=1cm,
    ]
    {
            &     & {$C^{uu}$} \\
            & {$C^u$} &   \\
      {$C_T$} &     & {$C^{ud/du}$} \\
            & {$C^d$} &   \\
            &     & {$C^{dd}$} \\
    };
    \draw[->] (tree-3-1) -- (tree-2-2) node [midway,above] {$P$};
    \draw[->] (tree-3-1) -- (tree-4-2) node [midway,below] {$(1-p)$};
    \draw[->] (tree-2-2) -- (tree-1-3) node [midway,above] {$P^2$};
    \draw[->] (tree-2-2) -- (tree-3-3) node [midway,below] {$(1-p)p$};
    \draw[->] (tree-4-2) -- (tree-3-3) node [midway,above] {$(1-p)p$};
    \draw[->] (tree-4-2) -- (tree-5-3) node [midway,below] {$(1-p)^2$};
  \end{tikzpicture}
```

If we assume dividends are paid as a % of $S_{t+\Delta}$. Then by
arbitrage: $$
\begin{bmatrix}
1 \\ S(t) \\ C(t)
\end{bmatrix} =
\begin{bmatrix}
B^u_{t+\Delta} & B^d_{t+\Delta} \\
S^u_{t+\Delta} + d_tS^u_{t+\Delta} & S^d_{t+\Delta} + d_tS^d_{t+\Delta} \\
C^u_{t+\Delta} & C^d_{t+\Delta} 
\end{bmatrix} 
\begin{bmatrix}
\psi_1 \\ \psi_2
\end{bmatrix}\\
E^Q[\frac{S_{t+\Delta}}{S}] = \frac{1+r\Delta}{1+d\Delta} \approx 1+(r-d)\Delta \\
E^Q[\frac{C_{t+\Delta}}{C}] = 1+r\Delta
$$ Under risk-neutral expectations. S grows below the risk free rate at
r-d while the call option is expected to return r. For foreign
currencies, replace d with the foreign savings interest rate rf.

## Risk Neutral Pricing Summary

-   Obtain a model to track dynamics of underlying
-   Calculate how derivative asset price relates to the price of the
    underlying asset ast expiration or other boundaries
-   Obtain risk adjusted probabilities
-   Calculate expected payoffs of derivative at expiration using risk
    adjusted probabilities
-   Discount this expectation using the risk-free return.

## Exotic Options {#Options}


### Lookback Options

Floating lookback call option payoff: $\max{(S_T - S_{min}, 0)}$

Fixed lookback call option payoff: $\max{(S_{max}-kK,0)}$

### Ladder Options

Has several thresholds, such that if the underlying price reaches these
thresholds, the return of the option is locked in.

### Trigger or Knock-in Options

A down-and-in option gives its holder a European option if the spot
price falls below a barrier during the life of the option. If the
barrier is not reached, the option expires with some rebate as a payoff.
The up-and-in option is similar but with the spot price going above the
barrier price.

### Knock-out Options

Kncok-out options are European options that expire immediately if the
underlying asset price falls below a barrier during the life of the
option The option pays a rebate if the barrier is reached. Also called
down and out. There are also up-and-out variations.

### Other Exotics

1.  *Basket options*, which are derivatives where the underlying asset
    is a basket of various financial instruments. Such basket dampen the
    volatility of the individual securities. Basket options become more
    afforable in the case of emerging market derivatives.
2.  *Multi-asset options* have payoffs depending on the underlying price
    of more than one asset

| Example Payoff          | $F(S_{1T}, S_{2T}, T)$                             |
|:------------------------|:---------------------------------------------------|
| Multi-Asset Option      | $\max{[0, \max{(S_{1T},S_{2T})-K}]}$               |
| Spread Call             | $\max{[0, (S_{1T}-S_{2T})-K]}$                     |
| Portfolio Call          | $\max{[0, (\theta_1S_{1T}+\theta_2S_{2T})-K]}$[^1] |
| dual strike call option | $\max{[0, S_{1T}-K, S_{2T}-K]}$                    |

[^1]: $\theta_2 , \theta_2$ represents the weight of each asset in the
    portfolio.

3.  *Average or Asian options* have payoffs depending on the average
    price of the underlying asset over the lifetime of the option.

## Interest Rate Sensitive Securities

-   So far we've discussed tools for securities that:
-   Don't pay dividends
-   Have risk-free interest rates
-   Underlying volatility is constant.
-   Derivatives in financial markets are more complicated.
-   Only applies to European Options.

### Interest Rate Derivatives

**Interest Rate Forwards and Futures**:

-   $L_{t_i}$ (Libor rates, average of loan quotes from selected banks in London) represents the annualized simple interest
    rate on a loan that begins at time $t_i$ and ends $t_{i+1}$
-   For a forward: The buyer of a forward will receive a loan with sum
    $N$ at time $t_i$
-   The loan is then paid back at time $t_{i+1}$ with amount
    $N(1+F_t \delta)$
-   $F_t$ is the interest rate and $\delta$ is the days adjustment
    factor, equal to the number of days between $t_i$, and $t_{i+1}$ over 365

**Forward Rate Agreements (FRA)**: As discussed in an earlier section,
these instruments provide a more convenient way to hedge interest rate
risk.

-   The buyer of an FRA either receives or pays $N[F_t - L_{t_i}]\delta$
    in-arrears at time $t_{i+1}$. (i.e. contract is established at time 0, Value of FRA contract determined at time 1, and then payment happens at time 2)
-   The buyer receives when $L_{t_{i}} < F_t$
-   The buyer pays when $L_{t_{i}} > F_t$
-   For a Standard FRA, the settlement date happens at the set date, which means that the payoff amount must be discounted back to time 1. $\text{Receipt} = \frac{N(F_t - L_{t_1})\delta}{1+L_{t_1}\delta}$

**Caps and Floors**:

-   Caps hedge against rising interest rates
-   Floors do the same but for decreasing rates.
-   The buyer of a caplet will receive $N \max{[\delta (L_{t_i}-R_{cap}),0]}$ where $R_{cap}$ is the cap rate
-   Can be thought of as a put on bond prices or call on interest rate.


```{r, echo=FALSE}
caps_floors <- cbind(c( "Caplet", "Floorlet"),
                          c( "$N \\max{(\\delta(L_{t_i} - R_{\text{cap}}),0)}$", "$N \\max{(R_{\text{floor}} - \\delta(L_{t_i}),0)}$" ),
                          c("Increase", "Decrease")
                          )

kableExtra::kbl(caps_floors, col.names = c("Instrument", "Payoff", "Gain When Interest Rates"), format = "html") %>% kable_minimal("hover")
```

**Interest Rate Swaps**:

-   As discussed in a prior chapter, the interest rate swap functions as a more complex FRA.
-   Involves an exchange of cash flows generated by a fixed pre-set swap rate $\kappa$ against cash flows generated by floating Libor rates $L_{t_i}$
-   The swap rate $\kappa$ is set so that the time $t$ price of the swap contract is 0
-   Can be thought of as a sequence of FRAs

**Bond Options**

-   A call option written on a bond gives its holder the right, but not the obligation, to buy a bond with price $B_t$ at strike price K

**Swaptions**

-   Options on swap contracts
-   The buyer of a swaption will have the right but not the obligation to get a swap contract at some future time

# Math of Financial Derivatives

### Basics

$$
1 + 1 + \frac{1}{2!} + \frac{1}{3!} \dots = e \\
e^{x+y} = e^x \cdot e^y\\
\frac{d}{dx} e^{kx} = ke^{kx} 
$$

**Function of Bounded Variation**

$$
V_0 = \max{\sum^n_{i=1}|f(t_i) - f(t_{i-1})|} < \infty
$$

**Definition of a derivative**

$$ 
f'_x = \lim_{\Delta \rightarrow 0} \frac{f(x+\Delta) - f(x)}{\Delta} \\
f_x'' = \frac{f_x'(x+\Delta) - f_x'(x)}{\Delta} \approx \frac{f(x+\Delta) - 2f(x) + f(x-\Delta)}{\Delta^2}
$$

**First order approximations of functions**

$$ 
f(x+\Delta) \approx f(x) + f'_x(\Delta)
$$

**Chain Rule**

$$
y_t = f(g(t)) \text{   and   } x_t = g(t) \\
\frac{dy}{dt} = \frac{df(g(t))}{dg(t)} \cdot\frac{dg(t)}{dt} 
$$

**Probability Basics**

-   $\omega = \text{particular state of the world}$
-   $\Omega = \text{all possible states of the world}$
-   $I = \text{set of all possible events}$
-   $P(A) \geq 0$ for any $A \in I$
-   $\int_{A \in I} dP(A) = 1$
-   distribution function: $G(x) = P(X \leq x)$

**Moments**

-   Expectation first moment: $E(X) = \int_{-\infty}^{\infty} xf(x)dx$
-   Second moment is variance
-   Third moment is skew
-   Fourth moment is kurtosis

**Normal Distribution**

$$
X \sim N(\mu, \sigma^2)
$$

-   PDF:
    $f(x) = \frac{1}{\sigma \sqrt{2\pi}}e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$
-   $E(X) = \mu$
-   $V(X) = \sigma^2$
-   skew = $E((X - E(X))^3) = 0$
-   kurtosis = $E[\left(\frac{x-\mu}{\sigma}\right)^4]$

### Intermediate

**Riemann Integral**

Given max $|t_i - t_{i+1}| \rightarrow 0$, the Riemann integral will be
defined by the limit

$$
\sum_{i=1}^n f\left(\frac{t_i + t_{i-1}}{2}\right)(t_i + t_{i-1}) \rightarrow \int_0^T f(s)ds
$$

**Integration by Parts**

$$
d(f(t) \cdot h(t)) = f'(t)h(t) + f(t)h'(t) \Rightarrow f(T)h(T) -f(0)h(0) - \int_0^Tf(t)h'(t)dt
$$

**Partial vs Total Differentials**

Let $C_t = F(S_t,t)$

$$
\frac{\partial F(S_t,t)}{\partial S_t} = F_s : \text{Partial Derivative} \\
df = F_sdS_t + F_tdt : \text{Total Differential}
$$

**1D Taylor Series**

Let f(x) be an infinitely differentiable function of $x \in \bR$.
Expanding around point $x_0$

$$
f(x) = f(x_0) + f'_x(x_0)(x-x_0) + \frac{1}{2}f''_x(x_0)(x-x_0)^2 + \dots = \sum_{n=0}^\infty \frac{f^n_x(x_0)}{n!}(x-x_0)^n
$$

**Ordinary Differential Equations**

$$
dB_t = -r_tB_tdt \Rightarrow B_t = e^{\int_0^tr_udu}
$$

**Conditional Expectation**

-   $I_{t_0} \subseteq I_{t_1} \subseteq \dots \subseteq I_{t_k}$
-   A family $I_t$ satisfying the above property is called filtration.
-   $I_t$ contains the information that we condition on
-   $P(|X-X_0| \leq \frac{dx}{2} \approx f(x_0)dx$
-   $E(S_t | I_u) = \int_{-\infty}^\infty S_t f(S_t | I_u) dS_t, u < t$
-   Linearity property: $E[S_t + F(t)] = E[S_t] + E[F(t)]$

**Moments**

-   Let $Y_t \sim N(\mu t, \sigma^2 t)$
-   Recall the definition of a moment generating function
    $M(\lambda) = E(e^{\lambda Y_t})$
-   Given the definition of $Y_t$ the moment generating function of
    $Y_t$ can be explicitly calculated:
    $M(\lambda) = e^{\lambda \mu t + \frac{1}{2}\sigma^2 t\lambda^2}$
-   Assume that $S_t = S_0 e^{Y_t}$. This assumes $S_t$ is lognormally
    distributed
-   Then the following conditional expectation follows from the moment
    generating function $M(1)$ over the time period $t - u$

$$
E[S_t | S_u, u < t] = S_u e^{\mu (t-u) + \frac{1}{2} \sigma^2 (t-u)}
$$

The following are key moments for a lognormal random variable L:

$$
E(L) = e^{\mu + \frac{1}{2} \sigma ^ 2} \\
V(L) = [e^{\sigma^2} - 1] e^{2\mu + \sigma^2} \\
E(L^n) = e^{n \mu + \frac{n^2 \sigma^2}{2}}
$$

# Pricing Methods

1.  Arbitrage on financial assets which are martingales. Known as the
    method of equivalent martingales.
2.  Construct a risk-free portfolio to obtain a partial-differential
    equation (PDE) that is implied by the lack of arbitrage
    opportunities.
    -   Gold Example:
        -   Portfolio 1: Borrow money, buy gold, hold until expiration.
            Pay back money at time T
        -   Portfolio 2: Buy forward contract (cash exchanged at T to
            receive gold).
        -   $F(S_t,t) = \underbrace{S_te^{r_t(T-t)}}_{\text{loan repayment}} + \underbrace{c(T-t)}_{\text{storage costs}}$
    -   Delta Hedging Example:
        -   Construct a risk free portfolio by combining loans, a long
            call, and short $\Delta$ shares. (requires continuous delta
            hedging)
        -   $\frac{\partial F(S_t,t)}{\partial S_t} = F(s) = \Delta$

**Binomial Models**

-   $Pr(\Delta F(t) = a \sqrt{\Delta} = p)$
-   Assume F(t) is the real time price, and that over a small time
    interval there is either an uptick or downtick.
-   $E[\Delta F(t) = ap\sqrt{\Delta} - a(1-p)\sqrt{\Delta} ]$
-   $V[\Delta F(t)] = 4a^2\Delta p(1-p)$

**Poisson Distribution**

-   Unpredictable occurrence at times t = 1,2,3.
-   Jumps are assumed to be independent and of the same size.
-   In a given small interval, the probability of observing more than 1
    jump is negligible.
-   Denote $N_t$ as number of jumps up to time t.
-   Let $\Delta$ be a small time interval.
-   $\lambda$ = positive constant called intensity
-   $\Delta N_t = 1$ if there is a jump, otherwise it's 0
-   $P(\Delta N_t =1) \approx \lambda N, P(\Delta N_t = 0) \approx 1-\lambda N, P(\Delta N_t = n) = \frac{e^{-\lambda\Delta}(\lambda \Delta)^N}{n!}$

**Markov Processes**

A discrete time process ${X_1, \dots ,X_t}$ with joint probability
distribution function $F(x_1,\dots, x_t)$ is said to be a Markov Process
iff:

$$
Pr(X_{t+s} \leq X_{t+s} | X_1, \dots, X_t) = Pr(X_{t+s} \leq X_{t+s} |  X_t)
$$

Where s \> 0, $P(. | I_t)$ is the probability conditional on the
information set $I_t$. Basically saying that the probability of the next
state is only conditional on the state prior to the event.

## Convergence

**Almost Sure Convergence**

A random variable $X_n$ converges to X almost surely if, for arbitrary
$\delta > 0$

$$
P(|\lim_{n \rightarrow \infty} X_n - X| > \delta ) = 0
$$

**Mean Square Convergence**

Let $X_0, X_1, X_2 \dots ,X_n$ be a sequence of random variables. Then
$X_n$ is said to converge to X in mean square if:

$$
\lim_{n \rightarrow \infty} E[(X_n - X)^2] =0
$$

**Weak Convergence**

$X_n$ = a random variable indexed by in with probability $P_n \cdot X_n$
converges to X weakly if $\lim_{n \rightarrow \infty} P_n = P$. Where P
is the probability distribution of X if:

$$
E^{P_n}[F(X_n)] \rightarrow E^{P}[F(X)]
$$

## Martingales

If the value of $S_t$ is included in the information set $I_t$ at t \>
0, then it is said that $\lbrace S_t, t \in [0,t]\rbrace$ is *adapted*
to $\lbrace I_t, t \in [0,\infty]\rbrace$. We can say that
$E_t[S_t] = E(S_t|I_t)$ forecasts $S_t$ given information at $I_t$.

A process $\lbrace S_t, t \in [0,\infty]\rbrace$ is a martingale w.r.t
the family of information sets $I_t$ and w.r.t probability P if for all
t \> 0, the following 3 properties are satisfied

1.  $S_t$ is known given $I_t$ ($S_t$ is $I_t$ adapted).
2.  $E^P|S_t| < \infty$ : probabilities are finite.
3.  $E^P[S_t] = S_t$ for all t \< T

Martingales satisfies the important property that it is not expected to
grow or shrink over time (the expectation = current value)
$E_t[S_{t+u} - S_t] = 0$

-   A **submartingale** is expected to grow over time or stay the same
    level.
    -   $E^P[X_{t+s} | I_t] \geq X_t$
-   A **supermartingale** is expected to shrink over time or stay the
    same level.
    -   $E^P[X_{t+s} | I_t] \leq X_t$

To convert to a martingale:

-   Change to a different probability measure.
-   Apply scaling factor (multiplicative/additive)

<!-- Suppose at time t one has information $I_t$. For a random variable $X_t$ -->

<!-- such that s \> 0 satisfies the equality: -->

<!-- -   $E^P[X_{t+s} | I_t] = X_t$ - This is a martingale with respect to -->

<!--     probability measure P -->

<!-- -   $E^P[X_{t+s} | I_t] \geq X_t$ - This is a submartingale with respect -->

<!--     to probability measure P -->

