---
title: "Exam QFI"
author: "Steven Cheng"
date: "2023-05-03"
output:
  html_document:
    theme: simplex
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
classoption:
- twocolumn
# header-includes:
#   - \usepackage{tikz}
#   - \ifdefined\HCode
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("kableExtra")
library("tidyverse")
  
```

\newpage

<span style="display: none;"> $$
\newcommand{\Lagr}{\mathcal{L}}
\newcommand{\bP}{\mathbb{P}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\bQ}{\mathbb{Q}}
\newcommand{\bE}{\mathbb{E}}
\Lagr \bP \bR \bQ \bE 
$$ <span style="display: none;">


# A Note About Notations


<span style="color: red;">Consider moving this as a
footnote<span style="color: red;">

In the book:

-   $dF_t$ represents the total change in the derivative price
    $F(S_t, t)$.
-   $F_t$ represents the partial derivative of $F(S_t, t)$ with respect
    to t.
-   In the notes, I attempt to make things more clear by denoting the
    the first order derivative with respect to t written as $F'_t$ and
    second order as $F''_t$
    
-   $B(t,T)$: price at time t<T of a discount bond maturing at time T
-   $r-t$: instantaneous spot rate on riskless borrowing
-   $R(t,T)$: yield of the discount bond $B(t,T)$
-   $F(t,T,U)$: forward rate at time t over the period T to U
-   $F(t,s)$: instantaneous forward rate contracted at time t over the interval s to s+ds
-   $(t-T)$: Often times represents the time to maturity and in some forumalas substituded to be $\tau$

<span style="color: red;">review chapter 5 for forward swap rates <span style ="color: red;">


# Financial Derivatives {#FD}

**Financial Derivatives** are securities that derive their value from
cash markets. In Ingersoll's definition, a financial contract is a
derivative security.

**Cash and Carry Markets**: Gold, silver, T-bonds. The idea is to borrow
cash, buy asset, and hold until expiration.

**Forward Contract**: obligation to buy/sell an asset at a specified
price at a future date. if cash price is higher at expiration then
forward, and long position make a profit.

```{r, echo = FALSE}
ForwardContracts <- cbind(c( "Mark to Market", "Settlement", "Margin System", "Trading", "Types", "Flexibility", "Liquidity", "Counterparty Risk"),
                          c( "Yes", "Daily", "Yes", "Exchanges", "Standardized" , "Less", "More", "Less"),
                          c("No", "Upon Maturity", "No", "OTC", "Custom", "More", "Less", "More")
                          )

kableExtra::kbl(ForwardContracts, col.names = c("Characteristic", "Futures", "Forwards"), format = "html") %>% kable_minimal("hover")


```

**Repurchase Agreement (REPO)**: A transaction in which one party sells
a security in return for cash, with an agreement to repurchase the
security at an agreed upon price at a future date.

[**Options**](#Options):

-   European: The right to buy $S_t$ at strike k. This right may be
    excercised at time T. Where $S_T <K$ is out the money and $S_T > K$
    is in the money. $$C_T = \max(S_T -K)$$

**Swaps**: Exchange of one set of cashflows for another. Swaps are
simultaneous and can involve curruncies, interst rates, etc. Swaps can
be priced through forwards. Cancelable swaps allow the right to cancel
swap contracts.

**Arbitrage** : Taking simultaneous positions so that it guarantees a
riskless profit higher than the risk-free rate. Let $d_{ij}$ be the
number of units of account paid by one unit of security i in state j.
Rows represent asset, columns represent state (interest, conflict,
recession, etc...)


::: columns
::: {.column width="20%"}
:::

::: {.column width="30%"}
$$
\text{Asset Values} = S_T = 
\begin{bmatrix} 
S_1(t) \\
S_2(t) \\
\vdots \\
S_n(t)
\end{bmatrix}
$$
:::

::: {.column width="30%"}
$$
D_t = \begin{bmatrix}
d_{11} & d_{12} & \cdots & d_{1k} \\
d_{21} & d_{22} & & d_{2k} \\
\vdots & & \ddots & \vdots\\
d_{i1} & d_{i2} & \cdots & d_{ik}
\end{bmatrix}
$$
:::

::: {.column width="20%"}
:::
:::

$$
\text{example of a 2 state portfolio with cash, security, and call}\\ \begin{bmatrix}
(1+r\Delta) & (1+r\Delta) \\
S_1(t + \Delta) & S_2(t + \Delta) \\
C_1(t + \Delta) & C_2(t + \Delta) 
\end{bmatrix}
$$

No arbitrage possibility exists iff:

$$
\begin{bmatrix}
1 \\ S(t) \\ C(t)
\end{bmatrix} 
=
\begin{bmatrix}
(1+r\Delta) & (1+r\Delta) \\
S_1(t + \Delta) & S_2(t + \Delta) \\
C_1(t + \Delta) & C_2(t + \Delta) 
\end{bmatrix} 
\begin{bmatrix}
\psi_1 \\ \psi_2
\end{bmatrix} 
$$

$\psi_n$ represents the risk neutral probability discounted at the risk
free rate and:

$$ \sum_i^n \psi_i = v = \frac{1}{1+r} \rightarrow \sum_i^n \psi_i \cdot (1+r) = \sum_i^n  \bQ_i  = 1 $$

|       **Binomial Tree Pricing**: A method of valuing options. Steps to determine option value is to first determine the value at expiration, then work backwards each step until the current price is obtained.

$$
C_T = \frac{1}{1+r}[Q_{up} C_{t+\Delta}^{up} + Q_{down} C_{t+\Delta}^{down}]
$$

```{tikz, BinomialTree, fig.ext = 'png', cache=TRUE, echo = FALSE, fig.dim = c(5, 4), fig.align="center"}
  \begin{tikzpicture}[>=stealth,sloped]
    \matrix (tree) [%
      matrix of nodes,
      minimum size=1cm,
      column sep=3.5cm,
      row sep=1cm,
    ]
    {
            &     & {$C^{uu}$} \\
            & {$C^u$} &   \\
      {$C_T$} &     & {$C^{ud/du}$} \\
            & {$C^d$} &   \\
            &     & {$C^{dd}$} \\
    };
    \draw[->] (tree-3-1) -- (tree-2-2) node [midway,above] {$P$};
    \draw[->] (tree-3-1) -- (tree-4-2) node [midway,below] {$(1-p)$};
    \draw[->] (tree-2-2) -- (tree-1-3) node [midway,above] {$P^2$};
    \draw[->] (tree-2-2) -- (tree-3-3) node [midway,below] {$(1-p)p$};
    \draw[->] (tree-4-2) -- (tree-3-3) node [midway,above] {$(1-p)p$};
    \draw[->] (tree-4-2) -- (tree-5-3) node [midway,below] {$(1-p)^2$};
  \end{tikzpicture}
```

If we assume dividends are paid as a % of $S_{t+\Delta}$. Then by
arbitrage: $$
\begin{bmatrix}
1 \\ S(t) \\ C(t)
\end{bmatrix} =
\begin{bmatrix}
B^u_{t+\Delta} & B^d_{t+\Delta} \\
S^u_{t+\Delta} + d_tS^u_{t+\Delta} & S^d_{t+\Delta} + d_tS^d_{t+\Delta} \\
C^u_{t+\Delta} & C^d_{t+\Delta} 
\end{bmatrix} 
\begin{bmatrix}
\psi_1 \\ \psi_2
\end{bmatrix}\\
E^Q[\frac{S_{t+\Delta}}{S}] = \frac{1+r\Delta}{1+d\Delta} \approx 1+(r-d)\Delta \\
E^Q[\frac{C_{t+\Delta}}{C}] = 1+r\Delta
$$ Under risk-neutral expectations. S grows below the risk free rate at
r-d while the call option is expected to return r. For foreign
currencies, replace d with the foreign savings interest rate rf.

## Risk Neutral Pricing Summary

-   Obtain a model to track dynamics of underlying
-   Calculate how derivative asset price relates to the price of the
    underlying asset ast expiration or other boundaries
-   Obtain risk adjusted probabilities
-   Calculate expected payoffs of derivative at expiration using risk
    adjusted probabilities
-   Discount this expectation using the risk-free return.

## Exotic Options {#Options}


### Lookback Options

Floating lookback call option payoff: $\max{(S_T - S_{min}, 0)}$

Fixed lookback call option payoff: $\max{(S_{max}-kK,0)}$

### Ladder Options

Has several thresholds, such that if the underlying price reaches these
thresholds, the return of the option is locked in.

### Trigger or Knock-in Options

A down-and-in option gives its holder a European option if the spot
price falls below a barrier during the life of the option. If the
barrier is not reached, the option expires with some rebate as a payoff.
The up-and-in option is similar but with the spot price going above the
barrier price.

### Knock-out Options

Kncok-out options are European options that expire immediately if the
underlying asset price falls below a barrier during the life of the
option The option pays a rebate if the barrier is reached. Also called
down and out. There are also up-and-out variations.

### Other Exotics

1.  *Basket options*, which are derivatives where the underlying asset
    is a basket of various financial instruments. Such basket dampen the
    volatility of the individual securities. Basket options become more
    afforable in the case of emerging market derivatives.
2.  *Multi-asset options* have payoffs depending on the underlying price
    of more than one asset

| Example Payoff          | $F(S_{1T}, S_{2T}, T)$                             |
|:------------------------|:---------------------------------------------------|
| Multi-Asset Option      | $\max{[0, \max{(S_{1T},S_{2T})-K}]}$               |
| Spread Call             | $\max{[0, (S_{1T}-S_{2T})-K]}$                     |
| Portfolio Call          | $\max{[0, (\theta_1S_{1T}+\theta_2S_{2T})-K]}$[^1] |
| dual strike call option | $\max{[0, S_{1T}-K, S_{2T}-K]}$                    |

[^1]: $\theta_2 , \theta_2$ represents the weight of each asset in the
    portfolio.

3.  *Average or Asian options* have payoffs depending on the average
    price of the underlying asset over the lifetime of the option.

## Interest Rate Sensitive Securities

-   So far we've discussed tools for securities that:
-   Don't pay dividends
-   Have risk-free interest rates
-   Underlying volatility is constant.
-   Derivatives in financial markets are more complicated.
-   Only applies to European Options.

### Interest Rate Derivatives

**Interest Rate Forwards and Futures**:

-   $L_{t_i}$ (Libor rates, average of loan quotes from selected banks in London) represents the annualized simple interest
    rate on a loan that begins at time $t_i$ and ends $t_{i+1}$
-   For a forward: The buyer of a forward will receive a loan with sum
    $N$ at time $t_i$
-   The loan is then paid back at time $t_{i+1}$ with amount
    $N(1+F_t \delta)$
-   $F_t$ is the interest rate and $\delta$ is the days adjustment
    factor, equal to the number of days between $t_i$, and $t_{i+1}$ over 365

**Forward Rate Agreements (FRA)**: As discussed in an earlier section,
these instruments provide a more convenient way to hedge interest rate
risk.

-   The buyer of an FRA either receives or pays $N[F_t - L_{t_i}]\delta$
    in-arrears at time $t_{i+1}$. (i.e. contract is established at time 0, Value of FRA contract determined at time 1, and then payment happens at time 2)
-   The buyer receives when $L_{t_{i}} < F_t$
-   The buyer pays when $L_{t_{i}} > F_t$
-   For a Standard FRA, the settlement date happens at the set date, which means that the payoff amount must be discounted back to time 1. $\text{Receipt} = \frac{N(F_t - L_{t_1})\delta}{1+L_{t_1}\delta}$

**Caps and Floors**:

-   Caps hedge against rising interest rates
-   Floors do the same but for decreasing rates.
-   The buyer of a caplet will receive $N \max{[\delta (L_{t_i}-R_{cap}),0]}$ where $R_{cap}$ is the cap rate
-   Can be thought of as a put on bond prices or call on interest rate.


```{r, echo=FALSE}
caps_floors <- cbind(c( "Caplet", "Floorlet"),
                          c( "$N \\max{(\\delta(L_{t_i} - R_{\text{cap}}),0)}$", "$N \\max{(R_{\text{floor}} - \\delta(L_{t_i}),0)}$" ),
                          c("Increase", "Decrease")
                          )

kableExtra::kbl(caps_floors, col.names = c("Instrument", "Payoff", "Gain When Interest Rates"), format = "html") %>% kable_minimal("hover")
```

**Interest Rate Swaps**:

-   As discussed in a prior chapter, the interest rate swap functions as a more complex FRA.
-   Involves an exchange of cash flows generated by a fixed pre-set swap rate $\kappa$ against cash flows generated by floating Libor rates $L_{t_i}$
-   The swap rate $\kappa$ is set so that the time $t$ price of the swap contract is 0
-   Can be thought of as a sequence of FRAs

**Bond Options**

-   A call option written on a bond gives its holder the right, but not the obligation, to buy a bond with price $B_t$ at strike price K

**Swaptions**

-   Options on swap contracts
-   The buyer of a swaption will have the right but not the obligation to get a swap contract at some future time



# Math of Financial Derivatives

### Basics

$$
1 + 1 + \frac{1}{2!} + \frac{1}{3!} \dots = e \\
e^{x+y} = e^x \cdot e^y\\
\frac{d}{dx} e^{kx} = ke^{kx} 
$$

**Function of Bounded Variation**

$$
V_0 = \max{\sum^n_{i=1}|f(t_i) - f(t_{i-1})|} < \infty
$$

**Definition of a derivative**

$$ 
f'_x = \lim_{\Delta \rightarrow 0} \frac{f(x+\Delta) - f(x)}{\Delta} \\
f_x'' = \frac{f_x'(x+\Delta) - f_x'(x)}{\Delta} \approx \frac{f(x+\Delta) - 2f(x) + f(x-\Delta)}{\Delta^2}
$$

**First order approximations of functions**

$$ 
f(x+\Delta) \approx f(x) + f'_x(\Delta)
$$

**Chain Rule**

$$
y_t = f(g(t)) \text{   and   } x_t = g(t) \\
\frac{dy}{dt} = \frac{df(g(t))}{dg(t)} \cdot\frac{dg(t)}{dt} 
$$

**Probability Basics**

-   $\omega = \text{particular state of the world}$
-   $\Omega = \text{all possible states of the world}$
-   $I = \text{set of all possible events}$
-   $P(A) \geq 0$ for any $A \in I$
-   $\int_{A \in I} dP(A) = 1$
-   distribution function: $G(x) = P(X \leq x)$

**Moments**

-   Expectation first moment: $E(X) = \int_{-\infty}^{\infty} xf(x)dx$
-   Second moment is variance
-   Third moment is skew
-   Fourth moment is kurtosis

**Normal Distribution**

$$
X \sim N(\mu, \sigma^2)
$$

-   PDF:
    $f(x) = \frac{1}{\sigma \sqrt{2\pi}}e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$
-   $E(X) = \mu$
-   $V(X) = \sigma^2$
-   skew = $E((X - E(X))^3) = 0$
-   kurtosis = $E[\left(\frac{x-\mu}{\sigma}\right)^4]$

### Intermediate

**Riemann Integral**

Given max $|t_i - t_{i+1}| \rightarrow 0$, the Riemann integral will be
defined by the limit

$$
\sum_{i=1}^n f\left(\frac{t_i + t_{i-1}}{2}\right)(t_i + t_{i-1}) \rightarrow \int_0^T f(s)ds
$$

**Integration by Parts**

$$
d(f(t) \cdot h(t)) = f'(t)h(t) + f(t)h'(t) \Rightarrow f(T)h(T) -f(0)h(0) - \int_0^Tf(t)h'(t)dt
$$

**Partial vs Total Differentials**

Let $C_t = F(S_t,t)$

$$
\frac{\partial F(S_t,t)}{\partial S_t} = F_s : \text{Partial Derivative} \\
df = F_sdS_t + F_tdt : \text{Total Differential}
$$

**1D Taylor Series**

Let f(x) be an infinitely differentiable function of $x \in \bR$.
Expanding around point $x_0$

$$
f(x) = f(x_0) + f'_(x_0)(x-x_0) + \frac{1}{2}f''_x(x_0)(x-x_0)^2 + \dots = \sum_{n=0}^\infty \frac{f^n_x(x_0)}{n!}(x-x_0)^n
$$

**Ordinary Differential Equations**

$$
dB_t = -r_tB_tdt \Rightarrow B_t = e^{\int_0^tr_udu}
$$

**Conditional Expectation**

-   $I_{t_0} \subseteq I_{t_1} \subseteq \dots \subseteq I_{t_k}$
-   A family $I_t$ satisfying the above property is called filtration.
-   $I_t$ contains the information that we condition on
-   $P(|X-X_0| \leq \frac{dx}{2} \approx f(x_0)dx$
-   $E(S_t | I_u) = \int_{-\infty}^\infty S_t f(S_t | I_u) dS_t, u < t$
-   Linearity property: $E[S_t + F(t)] = E[S_t] + E[F(t)]$

**Moments**

-   Let $Y_t \sim N(\mu t, \sigma^2 t)$
-   Recall the definition of a moment generating function
    $M(\lambda) = E(e^{\lambda Y_t})$
-   Given the definition of $Y_t$ the moment generating function of
    $Y_t$ can be explicitly calculated:
    $M(\lambda) = e^{\lambda \mu t + \frac{1}{2}\sigma^2 t\lambda^2}$
-   Assume that $S_t = S_0 e^{Y_t}$. This assumes $S_t$ is lognormally
    distributed
-   Then the following conditional expectation follows from the moment
    generating function $M(1)$ over the time period $t - u$

$$
E[S_t | S_u, u < t] = S_u e^{\mu (t-u) + \frac{1}{2} \sigma^2 (t-u)}
$$

The following are key moments for a lognormal random variable L:

$$
E(L) = e^{\mu + \frac{1}{2} \sigma ^ 2} \\
V(L) = [e^{\sigma^2} - 1] e^{2\mu + \sigma^2} \\
E(L^n) = e^{n \mu + \frac{n^2 \sigma^2}{2}}
$$

# Pricing Methods

1.  Arbitrage on financial assets which are martingales. Known as the
    method of equivalent martingales.
2.  Construct a risk-free portfolio to obtain a partial-differential
    equation (PDE) that is implied by the lack of arbitrage
    opportunities.
    -   Gold Example:
        -   Portfolio 1: Borrow money, buy gold, hold until expiration.
            Pay back money at time T
        -   Portfolio 2: Buy forward contract (cash exchanged at T to
            receive gold).
        -   $F(S_t,t) = \underbrace{S_te^{r_t(T-t)}}_{\text{loan repayment}} + \underbrace{c(T-t)}_{\text{storage costs}}$
    -   Delta Hedging Example:
        -   Construct a risk free portfolio by combining loans, a long
            call, and short $\Delta$ shares. (requires continuous delta
            hedging)
        -   $\frac{\partial F(S_t,t)}{\partial S_t} = F(s) = \Delta$

**Binomial Models**

-   $Pr(\Delta F(t) = a \sqrt{\Delta} = p)$
-   Assume F(t) is the real time price, and that over a small time
    interval there is either an uptick or downtick.
-   $E[\Delta F(t) = ap\sqrt{\Delta} - a(1-p)\sqrt{\Delta} ]$
-   $V[\Delta F(t)] = 4a^2\Delta p(1-p)$

**Poisson Distribution**

-   Unpredictable occurrence at times t = 1,2,3.
-   Jumps are assumed to be independent and of the same size.
-   In a given small interval, the probability of observing more than 1
    jump is negligible.
-   Denote $N_t$ as number of jumps up to time t.
-   Let $\Delta$ be a small time interval.
-   $\lambda$ = positive constant called intensity
-   $\Delta N_t = 1$ if there is a jump, otherwise it's 0
-   $P(\Delta N_t =1) \approx \lambda N, P(\Delta N_t = 0) \approx 1-\lambda N, P(\Delta N_t = n) = \frac{e^{-\lambda\Delta}(\lambda \Delta)^N}{n!}$

**Markov Processes**

A discrete time process ${X_1, \dots ,X_t}$ with joint probability
distribution function $F(x_1,\dots, x_t)$ is said to be a Markov Process
iff:

$$
Pr(X_{t+s} \leq X_{t+s} | X_1, \dots, X_t) = Pr(X_{t+s} \leq X_{t+s} |  X_t)
$$

Where s \> 0, $P(. | I_t)$ is the probability conditional on the
information set $I_t$. Basically saying that the probability of the next
state is only conditional on the state prior to the event.

## Convergence

**Almost Sure Convergence**

A random variable $X_n$ converges to X almost surely if, for arbitrary
$\delta > 0$

$$
P(|\lim_{n \rightarrow \infty} X_n - X| > \delta ) = 0
$$

**Mean Square Convergence**

Let $X_0, X_1, X_2 \dots ,X_n$ be a sequence of random variables. Then
$X_n$ is said to converge to X in mean square if:

$$
\lim_{n \rightarrow \infty} E[(X_n - X)^2] =0
$$

**Weak Convergence**

$X_n$ = a random variable indexed by in with probability $P_n \cdot X_n$
converges to X weakly if $\lim_{n \rightarrow \infty} P_n = P$. Where P
is the probability distribution of X if:

$$
E^{P_n}[F(X_n)] \rightarrow E^{P}[F(X)]
$$

## Martingales

If the value of $S_t$ is included in the information set $I_t$ at t \>
0, then it is said that $\lbrace S_t, t \in [0,t]\rbrace$ is *adapted*
to $\lbrace I_t, t \in [0,\infty]\rbrace$. We can say that
$E_t[S_t] = E(S_t|I_t)$ forecasts $S_t$ given information at $I_t$.

A process $\lbrace S_t, t \in [0,\infty]\rbrace$ is a martingale w.r.t
the family of information sets $I_t$ and w.r.t probability P if for all
t \> 0, the following 3 properties are satisfied

1.  $S_t$ is known given $I_t$ ($S_t$ is $I_t$ adapted).
2.  $E^P|S_t| < \infty$ : probabilities are finite.
3.  $E^P[S_t] = S_t$ for all t \< T

Martingales satisfies the important property that it is not expected to
grow or shrink over time (the expectation = current value)
$E_t[S_{t+u} - S_t] = 0$

-   A **submartingale** is expected to grow over time or stay the same
    level.
    -   $E^P[X_{t+s} | I_t] \geq X_t$
-   A **supermartingale** is expected to shrink over time or stay the
    same level.
    -   $E^P[X_{t+s} | I_t] \leq X_t$

To convert to a martingale:

-   Change to a different probability measure.
-   Apply scaling factor (multiplicative/additive)

<!-- Suppose at time t one has information $I_t$. For a random variable $X_t$ -->

<!-- such that s \> 0 satisfies the equality: -->

<!-- -   $E^P[X_{t+s} | I_t] = X_t$ - This is a martingale with respect to -->

<!--     probability measure P -->

<!-- -   $E^P[X_{t+s} | I_t] \geq X_t$ - This is a submartingale with respect -->

<!--     to probability measure P -->

# Brownian Motion and Wiener Processes

**Understanding Variance**

-   $V(X) \approx E(X^2)$ : Assume the interval is small enough where
    the expected movement is negligible.
-   $V(X) \ approx \sum_k p_k k^2 \propto h \text{ where } p_k = \Pr(X=k)$
-   k = size term and $p_k$ = probability
-   The probability times the size squared will be linear to h which
    leads to the following 3 cases:
    -   Probability proportional to h, size independent of h
    -   Probability independent of h, size propotional to $\sqrt{h}$
    -   Something more complicated and in between

**Brownian Bridge**

::: columns
::::: {.column width="\"50"}
$$
\text{For }i \in \lbrace1,2 \dots,n \rbrace \text{ define:} \\
X_i(t) = \begin{cases}
1 & \text{if } \in_i < t\\
0 & \text{otherwise}
\end{cases}\\
\text{Then let } S_n(t) = \frac{1}{\sqrt{n}}\sum^n_{i=1}(X_i(t)-t)
$$
:::::

::::: {.column width="50%"}
$\in_i$ assumed to have uniform distribution on (0,1)

In the limit as $n \rightarrow \infty$, this is called the Brownian
Bridge
:::::
:::

## Brownian Motion

Brownian motion is approximated by Taylor Series Approximation. Suppose
that x is a martingale process.

$$
f(x_0 - \Delta x) - f(x_0) = f'_x(\Delta x) + \frac{1}{2}f''_x(\Delta x) ^2 + \cdots
$$

Usually we ignore higher order terms and just look at first order
approximation for deterministic calculus. However we cannot ignore
higher order terms for stochastic calculus. Getting rid of second order
term in stochastic calculus means that we assume the underlying random
variable has:

$$
\matrix{
V(\Delta x) = E[(\Delta x)^2] - E[(\Delta x)]^2 = E[(\Delta x)^2] = 0 & \text{Wrong for stochastic setting}
}
$$

**Definition of Brownian Motion**

A random process $B_t, t \in [0,T]$ is a brownian motion if:

1.  $B_0 = 0$
2.  $B_t$ has stationary, independent increments
3.  $B_t$ is continuous in t.
4.  $B_t - B_s \sim N(0,|t-s|)$

## Wiener Process

**Definition of a Wiener Process**

A Weiner process, $W_t$, relative to a family of information sets
${I_t}$, is a stochastic process such that:

1.  $W_t$ is a square integrable martingale
2.  $W_0 = 0$
3.  $E[(W_t - W_s)^2] = t-s, s\leq t$
4.  The trajectories of $W_t$ are continuous over t.

**Properties of Wiener Process**

1.  $W_t$ has uncorrelated increments )note every martingale has
    unrelated increments).
2.  $W_t$ has 0 mean. (Since it starts at 0, all increments are expected
    to have 0 movement).
3.  $W_t$ has variance t
4.  The process is continuous; in infinitesimal intervals, the movements
    of $W_t$ are infinitesimal.

**Brownian Motion vs Wiener Process**

-   Any Wiener process $W_t$ relative to a family $I_t$ is a Brownian
    motion process
-   THis is the Levy theorem, and it states that there is no difference
    between a Wiener process and Brownian motion

**Stock Fluctuations From Unpredictable Information**

-   Assume there is no drift and that when news comes out the stock
    price moves by $\sigma_t dW_t$

$$
E_t[\int_t^{t+T}\sigma_u dW_u] = 0
$$

## Poisson Process

$$
dN_t = \cases{
 1 & with probability $\lambda$ dt \\
 0 & with probability $1-\lambda$ dt
}
$$

-   The size of poisson outcome does not depend on the size of the time
    interval dt.
-   The probability of each outcome does depend on the size of the time
    interval dt
-   $N_t$ is not a martingale since $E(dN_t) = \lambda dt \neq 0$

Can convert to a martingale transforming $\mu_t = N_t - \lambda t$,
$E[M_t] = 0$, As $dt \rightarrow 0$, the probability of a Poisson jump
goes to 0.

Wiener Integral : $\int_{t_0}^T f(W_t)dW_t$ \* Very irregular snce there
are uncountably many changes as $t \rightarrow 0$ \* Unbounded variation
\* Riemann-Stieltjes definition does not apply

Poisson Integral : $\int_{t_0}^T f(M_t)dM_t$ \* As \$ t \rightarrow 0\$,
chance of jump goes to 0 \* Bounded variation \* Riemann_Stieltjes
definition applies.

## Characterizing Rare and Normal Events

Assume $\Delta W_k$ can only take on a finite number of possible values.

$$
\sigma_k \Delta W_k = \begin{cases}
w_1 & \text{probability } p_1 \\
w_2 & p_2 \\
\vdots \\
w_n & p_n
\end{cases}
$$ $$
E[(\sigma_k \Delta w_k)^2] = \sigma^2_k h
$$

$$
V[\sigma_k \Delta w_k] = \sigma^2_kh = \sum^m_{i=1} p_i w_i^2 \rightarrow p_i w_i^2 = c_ih
$$ Assume we can specify $w_i$ and $p_i$ in the following forms:
<span style="color: red;">[FLAG] what is w?<span style="color: red;">

$$
w_i(h)  = \bar w_i h^{r_i} \\
p_i(h) = \bar p_i h^{q_i} \\ 
$$

$$
\text{Then: } p_i w_i^2 = c_i h - \underbrace{p_i w_i^2}_{c_i} h^{q_i}h^{2r_i} = q_i + 2r_i = 1 \text{ where } \begin{matrix} 0 \leq r_i \leq \frac{1}{2} \\ 0 \leq q_i  \leq 1 \end{matrix}
$$

**Summary Table for Characterizing Rare and Normal Events**

| Event  |     $r_i$     | $q_i$ |  Limiting Size  | Limiting Prob  | Time Independence |
|:----------:|:----------:|:----------:|:----------:|:----------:|:----------:|
| Normal | $\frac{1}{2}$ |   0   |        0        | $\bar p_i > 0$ |    probability    |
|  Rare  |       0       |   1   | $\bar w_i \> 0$ |       0        |       size        |

# Stochastic Calculus

## Stochastic Derivative

Modify definition of the derivative slightly to include extra term:

$$
\lim_{x \rightarrow 0} \frac{f(x + \Delta x) - f(x)}{\Delta x} \approx f'_x + \frac{1}{2}f''_x\Delta x
$$

### Framework

-   $S_k = S(kh)$
-   Divide an interval length T into n intervals of length h:

$$
 \sigma_k \Delta W_k = \underbrace{[S_k - S_{k-1}]}_\text{innovation} - \underbrace{E_{k-1}[S_k - S_{k-1}]}_\text{expected movement}
$$

-   Key property $E[\Delta W_k] = 0$ for all k.
-   $E_{k-1}[\Delta W_k] = W_{k-1}$
-   Define the martingale $W_k = \sum_{i=1}^k \Delta W_i$

#### Analyzing Innovation Term

Define (note that $V^k$ represents variance):

$$
V^k = E_0[(\Delta W_k)^2] \\
V = E_0[\sum_{k=1}^n \Delta W_k]^2 = \sum_{k=1}^n E_0[( \Delta W_k)^2] = \sum_{k=1}^nV^k \\
V_\text{max} = \max[V^k, k=1,2,\dots,n]
$$

Let there be n intervals of length h and T = nh. Let
$\Delta W_k = W((k+1)h) - W(kh)$.

**Assumptions**

-   $0 < V < \infty$ : variation is non-zero and finite
-   $0 < \frac{V^k}{V^\max} \leq 1$ : no subinterval dominates the
    randomness.

As long as the assumptions are met, then the variance of $\Delta W_k$ is
proportional to h.

$$
E_[(\Delta W_k)^2] = h
$$

giving us the approximation: $\Delta W_k \approx \sqrt{h}$. If we had
tried to apply this to a traditional deterministic method we get:

$$
\lim_{h \rightarrow 0^+} \frac{\Delta W_k}{h} \approx \lim_{h \rightarrow 0^+} \frac{\sqrt{h}}{h} = {\frac{1}{\sqrt{h}} \rightarrow \infty} \text{ (Impossible!)}
$$

rearranging:

$$
\sigma_k \Delta W_k = [S_k - S_{k-1}] - E_{k-1}[S_k - S_{k-1}] \\ 
\downarrow \\
dS_t = \underbrace{a (S_t,t)dt}_\text{drift} + \underbrace{b(S_t, t) dW_t}_\text{diffusion}
$$ **Takeaway**

$dW_t$ is irregular and its derivative does not exist in the traditional
sense.

$$
\matrix{E[(\Delta W_k])^2]=h & E[\Delta W_k] = 0 & V[\Delta W_k] = h}
$$ The distribution of $W_{t+h} - W$ conditioned on the information set
$I_t$ is normally distributed with mean 0 and variance h.

### Ito's

#### Ito Integral <span style="color: red;">[FLAG] need to review from book?<span style="color: red;">

$$
\int_0^t \sigma(S_u,u)dW_u
$$

1.  $\lim_{sup_i |t_{i r} - t_i| \rightarrow 0} \sum^{n-1}_{i=0} g(X_{t_{i+1}})[F(X_{t_{i+1}}) - F(X_{t_i})] = \int_0^T g(X_T)dF(X_t)$
2.  $\lim_{n \rightarrow \infty} \sum^{n-1}_{k=1} \sigma (S_{k-1}, k)[w_k - w_{k-1}] = \int_0^T \sigma(S_u,u)dW_u$
3.  Used for constructing stochastic integrals involving Wiener Process.
    $$
    \lim_{n \rightarrow \infty} E [\sum^{n-1}_{k=1} \sigma (S_{k-1}, k)[w_k - w_{k-1}] -  \int_0^T \sigma(S_u,u)dW_u]^2 =0
    $$

**Ito Definition**

Consider the finite interval approximation of the stochastic
differential equation.

$$
S_k - S_{k-1}  = a (S_{k-1}, k)h + \sigma (S_{k-1}, k)[w_k - w_{k-1}]
$$

when $[w_k - w_{k-1}]$ is a standard Wiener process with 0 mean and
variance h.

-   $\sigma (S_t,t)$ is non-anticipative (independent of the future)
-   $\sigma (S_t,t)$ is non-explosive
    $E[\int_0^T \sigma(S_t,t)^2 dt] < 0$

Then the Ito Integral is the mean square limit

$$
\int_0^T \sigma (S_t,t) dW_t = \sum^n_{k=1} \sigma (S_k,k)[w_k - w_{k-1}]
$$

Additional notes:

-   The Ito Integral here is $\int_0^T \sigma (S_t,t) dW_t$
-   Ito is a random variable and a martingale
-   Ito Integral is a mean-square limit

**Stochastic vs Deterministic Integrals**

-   The notion of a limit involves random variables
-   Ito integral defined for non-anticipative functions only
-   Deterministic calculus uses actual paths followed by functions,
    stochastic integrals use stochastic equivalence.

#### Ito's Isometry

For any $X_t(w)$ which is square integrable, the following holds

$$
E[(\int_0^t X_u(w)dW_u(w))^2] = E[\int_0^t X_u(w)]^2 du
$$

-   *Existence*: If f is continuous and non-anticipating, Then the Ito
    Integral exists
-   *Martingale*: $E[\int_0^T f(W_t,t)dW_t] = 0$
-   *Additive*

Products

1.  $E[\int_0^T f(W_t,t)dW_t \int_0^T g(W_t,t)dW_t] = E[\int_0^T f(W_t,t)g(W_t,t)dW_t]$
2.  $E[\int_0^T f(W_t,t)dW_t \int_0^T f(W_t,t)dW_t] = E[\int_0^T f(W_t,t)^2 dW_t]$

#### Ito's Lemma

1.  Identify $F(S_t,t)$
2.  Identify $a_t$ and $\sigma_t$
    -   $dS_t = a_t dt + \sigma_t dW_t$
3.  Calculate Partial Derivatives
    -   $\frac{\partial F}{\partial S_t }$ , $\frac{\partial F}{\partial t}$
    , $\frac{\partial^2 F}{\partial S_t^2}$
4.  Plug and Chug
    -   $dF_t = [a_t \frac{\partial F}{\partial S_t} + \frac{\partial F}{\partial t} + \frac{1}{2}\sigma \frac{\partial^2 F}{\partial S_t^2}]dt + \frac{\partial F}{\partial S_t } \sigma_t dW_t$

### Common SDE's

**Linear:**

$$
\begin{aligned}
& dS_t = a(S_t,t)dt + \sigma (S_t,t) dW_t \\
&  a(S_t,t)  = \mu \\
&  \sigma (S_t,t) = \sigma
\end{aligned}
$$

-   This gives the linear SDE: $dS_t = \mu dt + \sigma dW_t$
-   linear trend $\mu$ with noise term scaled by $\sigma$

**Geometric Brownian Motion (GBM):**
$dS_t = \mu S_t dt + \sigma S_t dW_t$

-   This is an exponential trend $e^\mu$ with noise scaled by $\sigma$
-   Divide both sides by $S_t$ to get an intuition w/ percentages
-   $$
      \underbrace{\frac{dS_t}{S_t}}_{\text{ Actual % Change}} = \overbrace{\mu dt}^{\text{Expected % Change}} + \underbrace{\sigma dW_t}_{\text{Unexpected % Change}}
      $$
-   Equivalently written as
    $d \ln(S_t) = (\mu - \frac{1}{2}\sigma^2)dt = \sigma dW_t$

**Square Root Process:** $dS_t = \mu S_t dt + \sigma \sqrt{S_t} dW_t$

-   Reduces randomness by taking sqrt of the noise component

**Mean-Reverting Process:**
$dS_t = \lambda (\mu - S_t) dt + \sigma S_t dW_t$

-   \$\lambda \> 0 \$: rate of mean reversion
-   $\mu$ is the level of mean reversion
-   if $\mu < S_t$, drift term is negative so the process drifts to
    $\mu$ <span style="color: red;">[FLAG] review
    drift<span style="color: red;">
-   if $\mu > S_t$, drift term is positive, $S_t$ drifts back up to
    $\mu$

**Ornstein-Uhlenbeck Process:** $dS_t = -\mu S_t dt + \sigma dW_t$

-   $\mu > 0$ sets the speed of reversion
-   Mean reverting process around $S_t = 0$

**Summary**

| SDE Name           | SDE Form                                          |
|:-------------------|:--------------------------------------------------|
| Linear             | $dS_t = a(S_t,t)dt + \sigma (S_t,t) dW_t$         |
| GBM                | $dS_t = \mu S_t dt + \sigma S_t dW_t$             |
| Square Root        | $dS_t = \mu S_t dt + \sigma \sqrt{S_t} dW_t$      |
| Mean Reverting     | $dS_t = \lambda (\mu - S_t) dt + \sigma S_t dW_t$ |
| Ornstein-Uhlenbeck | $dS_t = -\mu S_t dt + \sigma dW_t$                |

## Stochastic Variance

$dS_t = \mu S_t dt + \sigma S_t dW_t \rightarrow d\sigma_t = \lambda(\sigma_0 - \sigma_t)dt + \alpha \sigma_t dW_t$

2 kinds of jump framework:

1.  Jump-diffusion models: Adds in a jump component to the SDE to our
    typical diffusion model.
2.  Variange-Gamma: perform a stochastic time change

### Jump-Diffusion

-   $\frac{d S_t}{S_t} = \mu dt + \sigma dW_t$
-   Jump occurs with probability $\lambda dt$
-   jump occurs of $dN_t = 1$ otherwise $dN_t =0$
-   The expected proportion jump size is $K = E(e^{J} -1)$
-   J \> 0 upward jump, J \< 0 downward jump
-   Jump-Diffusion Model cn be based on
    $\frac{d S_t}{S_t} = (\mu - \lambda K) dt + \sigma dW_t + (e^J -1) dN_t$

### Variance-Gamma (VG) Process

-   Performs a stochastic time change. Adjusting time period will adjust
    the volatility, so this is a stochastic variance model as well.
-   The stochastic time is $\gamma(t; 1,v)$ which is Gamma distributed
-   The process given the stochastic time is
    $b(t, \sigma, \theta) = \theta t + \sigma W_t$
-   The unconditional process is defined by:
    $X(t; \sigma, v, \theta) = b(\gamma(t; 1,v), \sigma, \theta) = \theta \gamma(t; 1,v) + \sigma W(\gamma(t; 1,v))$
-   If $t^* = \gamma(t; 1,v)$, then
    $X(t; \sigma, v, \theta) = \theta t^* + \sigma W(t^*)$

# Black-Scholes PDE Derivation

## Forming Risk-Free Portfolios

Suppose $\theta_1 \& \theta_2$ are the quantities of the derivative
instrument and the underlying security purchased, respective $P_t$ is
the risk free portfolio.

$$
\begin{aligned}
& dS'_t = a_tdt + \sigma_t dW_t \\
& dF'_t = F'_s a_t + 0.5F''_s \sigma^2_t + F_t]dt + F'_s \sigma_t dW_t
\end{aligned}
$$

$$
\begin{aligned}
P_t &= \theta_1 dF_t + \theta_2 dS_t \\
&= \theta_1 ([F'_s a_t + 0.5F''_s \sigma^2_t + F'_t]dt + F'_s \sigma_t dW_t) + \theta_2(a_tdt + \sigma_t dW_t)
\end{aligned}
$$

By setting $\theta_1 = 1$ and $\theta_2 = -F'_s$ (equivalent to delta
hedge) Then $dP_t = F'_t dt + 0.5 F''_s \sigma^2_t dt$. Because there is
no random noise then we must have that
$dP_t = rP_t dt = F'_t dt + 0.5 F''_s \sigma^2_t dt$. (r = risk free
rate)

Note that: $P_t = F(S_t,t) -F'_sS_t$

Combining, we get the **Black-Scholes PDE**

$$
r(F - F'_sS_t)dt = F'_t dt + 0.5 F''_S \sigma^2_t dt \\
\rightarrow -rF + rF'_sS_t + F'_t + 0.5F''_s \sigma^2(t) = 0
$$

Note that $\sigma_t = \sigma S_t$

## PDE Classification

**Linear and non-linear:** this deals with how partial derivatives are
aggregated

-   Linear: $F'_s + 0.2F'_t = 4$
-   non-linear $F'^2_s + F'^3_t = 4$

**Order:** this deals with the highest power of partial derivative
involved

-   First Order: $F'_s + 0.2 F'_t = 4$
-   Second Order: $F''_s + 0.2F''_t = 4, F'_{st} + F_t = 1$

$a_0 + a_1F'_t + a_2F'_s + a_3 F''_s + a_4 F''_t + a_5 F'_{st} = 0$

-   Elliptic PDE: $a^2_5 -4a_3a_4 < 0$
-   Parabolic PDE: $a^2_5 -4a_3a_4 = 0$
-   Hyperbolic PDE: $a^2_5 -4a_3a_4 > 0$

## Black-Scholes

1.  Assume that $dS_t = a(S_t,t)dt + \sigma (S_t,t)dW_t$
2.  Under Black-Scholes, we assume that
    -   $a(S_t,t) = \mu S_t$
    -   $\sigma(S_t,t) = \sigma S_t$
3.  The PDE and boundary conditions are given by:
    -   $-rF + rF'_s S_t + F'_t + 0.5 \sigma^2 F''_s s^2_t = 0$
    -   $F(T) = \max(S_T-K, 0)$ (Assumes we're looking at a call option)
4.  Solution to Black-Scholes given boundary conditions.
    -   $F(S_t,t) = S_t N(d_1) - Ke^{-r(T-t)}N(d_2)$
    -   $d_1 = \frac{ \ln{\frac{s_t}{k}}+ (r + 0.5 \sigma^2)(T-t)}{\sigma \sqrt{T-t}}$
    -   $d_2 = d_1 - \sigma \sqrt{T-t}$

Recall that $N(d_i), i = 1,2$ are the two integrals of the standard
normal density:
$N(d_i) = \int_{-\infty}^{d_i} \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2}x^2}dx$.

Solutions can be verified by taking all the partial derivatives in the
Black-Scholes PDE and show that the sum equals 0. Verify that the
boundary conditions work.

**Assumptions of Black-Scholes**

1.  The underlying asset is a stock
2.  The stock does not pay dividends
3.  The derivative asset is a European-style call option (i.e. cannot be
    exercised before the expiration date)
4.  The risk-free rate r is constant
5.  there are no indivisibilities or transaction costs such as
    commissions and bid-ask spreads.

### Geometric Look at Black-Scholes

-   Fix r, K, $\sigma$ and T. Assume no dividends
-   A call price surface will be defined as a three-dimensional graph
    $C_t = F(S_t, T-t)$ where the value of the call is determined by
    varying the strike and time to expiration.

### Partial-Integro Differential Equations (PIDE)

<span style="color: red;">I don't believe the exam will test extensively
on PIDE's. Hopefully just knowing what they are is enough and we won't
have to use them. I have no idea where the (r-q) term comes from in the
equation below either.<span style="color: red;">

Consider the Black-Scholes PDE:

$$
\frac{\partial F}{\partial t} + \underbrace{\frac{\sigma^2 S^2}{2}\frac{\partial^2F}{\partial S^2}}_{\text{Diffusion Term}} + (r-q)S \frac{\partial F}{\partial S} = rF(S,t)
$$

Pricing European options under the variance-gamma model involves first
conditioning on the random-time g and then simplifying using a
Black-Scholes type formula to solve for the conditional option value.

The levy measure for the variance gamma process given by: 
$$
\begin{matrix}
\underbrace{dv(y)}_{\text{levy measure under Q}} = k(y)dy & \text{Define } k(y) & 
\begin{cases} 
\text{if } y> 0 ,  k(y) = \frac{e^{-\lambda_p} \rho^y}{vy^{1+Y}} \\
\text{if } y< 0 ,  k(y) = \frac{e^{-\lambda_n} \rho^y}{v|y|^{1+Y}} \\
\lambda_p = \sqrt{\frac{\theta^2}{\sigma^4} + \frac{2}{\sigma^2 v}} - \frac{\theta}{\sigma^2} \\
\lambda_n = \sqrt{\frac{\theta^2}{\sigma^4} + \frac{2}{\sigma^2 v}} + \frac{\theta}{\sigma^2}
\end{cases}
\end{matrix}
$$

Behavior of Jumps Given Y.

| Threshold | Behavior Below                    | Behavior Above                  |
|:-----------------------|:-----------------------|:-----------------------|
| Y = 1     | Finite Variation                  | Infinite Variation              |
| Y = 0     | Finite Arrival Rate               | Infinite Arrival Rate           |
| Y = -1    | Activity concentrated away from 0 | Activity concentrated towards 0 |

The PIDE for Variance Gamma is:

$$
\int_{-\infty}^\infty [F(S_{t-} e^y, t) - F(S_{t-},t) - \frac{\partial V(S_{t-, t})}{\partial S} S_t (e^y - 1)]k(y) dy+ \frac{\partial V (S_t,t)}{\partial t} + (r-q) S_t \frac{\partial V (S_t,t)}{\partial S} - rF(S_t,t) = 0
$$

The first term of the PIDE is the complications due to the stochastic
time component. Back terms are similar to Black-Scholes. We can think of
$k(y)$ as the function that gives the behavior of the jumps as
illustrated in the table. Rewriting Black-Scholes as a PIDE (steps
involved here are tedious but involves substituting the second
derivative of F and S using chain rule, substituting $\sigma^2$ with the
levy density, applying an Taylor Expansion, dropping off 3rd order terms
from expansion, and simplifying):

$$
\underbrace{\int_{|y|\leq \epsilon} F(Se^y, t) - F(S_{t},t) - (e^y - 1)\frac{\partial F(S, t)}{\partial S} ]k(y) dy}_{\text{diffusion term}} + \frac{\partial F}{\partial t} + (r-q) S \frac{\partial F}{\partial S}= rF(S_t,t) 
$$ 

1. The diffisuion term is replaced by an integral in a narrow range
2. The PIDE is better for fitting smiles (related to implied vol which
is expected to be constant under the regular Black-Scholes PDE. PIDE's
allow the integral part of the diff eq. to add additional complexity in
the volatility component):

### Complications of Black-Scholes

1. Has to meet the assumptions discussed previously. No early-exercise possibility, and also the risk free interest rate is assumed to be constant.
2. Many interest sensitive securities have early exercise provisions
3. Assets with term structures or random expiration dates (such as with American Options or Barrier Derivatives)
4. Bonds converge in price to the face value as time approaches expiration, so the assumption of constant volatility is unrealistic.
5. Mark-to-market securities will also introduce complications since these cash flows will be discounted by stochastic discount factors.
6. Additional modifications necessary for bonds that pay coupons.
7. Girsanov correspondence is not obvious.
8. Does not take into account multiple underlying assets.


# Equivalent Martingale Measures

The idea is that we want to transform the probability measure $d \bP$ so
that the mean of a random process $z_t$ changes. The transformation
allows us to treat assets with positive "risk premium" to be treated as
if they were risk free.

**Method 1:** subtracting means: $\bar{z} =z - E[z]$

**Method 2:** using equivalent measures: Transform $\bP$ to $\bQ$ so
$d \bQ = \xi(z_t) d \bP(z_t)$. See [Girsanov Theorem](#Girsanov)

**Probability as a measure**

-   Use measures to quantify the probability that a random variable
    falls within a small interval around z
-   $\int_{-\infty}^{\infty} d \bP (z_t)=1$
-   $E(z_t) - \int_{-\infty}^\infty z_t d\bP(z_t)$
-   $V(z_t) = E[z_t - E(z_t)]^2 = \int_{-\infty}^\infty[z_t-E(z_t)]^2d\bP z_t$

Consider a normally distributed random variable $z_t$ with fixed t and
normal density function $d\bP(z_t)$:

$$
z_t \sim N(0,1) \\
d\bP = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}(z_t)^2} dz_t
$$ Next, we define the function
$\xi (z_t) = e^{z_t\mu - \frac{1}{2}\mu^2}$ that when we multiply
$\xi(z_t) d\bP(z_t)$ we obtain a new probability:

$$
[d\bP(z_t)][\xi(z_t)] = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}(z_t)^2 + \mu z_t - \frac{1}{2} \mu ^2}  dz_t \\
d\bQ(z_t) = \frac{1}{\sqrt{2\pi}}e^{-\frac{1}{2}(z_t - \mu)^2} dz_t \\
$$ 

Clearly $d\bQ(z_t)$ is a new probability measure defined as:

$$
d\bQ(z_t) = \xi(z_t) d\bP(z_t)
$$

## Radon-Nikodym Derivative

-   $\frac{d \bQ (z_t)}{d \bP(z_t)} = \xi(z_t)$
-   The function $\xi$ allows us to transform between $\bP$ and $\bQ$
-   If both $\bP$ and $\bQ$ have the same intervals of $dz_t$ with non
    zero probability, then the Radon-Nikodym Derivative exists

## Converting Asset Prices into Martingales

-   Assume that under $\bP, Y_t \sim N(\rho t, \sigma^2 t)$
-   Under the true probability distribution
    $\bP, E^\bP[e^{-rt} S_t | S_u, u < t]$ is likely not true since, in
    general, there is a risk premium
-   Switch to the risk-neutral probability $\bQ$, and then
    $E^\bQ[e^{-rt} S_t | S_u, u < t] = e^{-ru} S_u$

From a SDE viewpoint:

-   Let $F_t = S_0e^{Y_t}$ where $dY_t = \mu dt + \sigma dW_t$
-   $dF_t = [\mu \frac{\partial F}{\partial Y_t} + \frac{1}{2} \sigma^2_t \frac{\partial ^2 F}{\partial Y_t^2}]dt + \frac{\partial F}{\partial Y_t} \sigma_t dW_t$
-   All of the partial derivatives shown simply equal $F_t$
-   $dF_t = [\mu + \frac{1}{2} \sigma^2_t]F_t dt + \sigma_t F_t dW_t$
-   Using Ito's Lemma gives that the true probability $\bP$ has a drift
    coefficient of $(\mu + \frac{1}{2})S_t$
-   The true probability SDE uses the Wiener process $W_t$
-   Setting $\mu = r - \frac{1}{2} \sigma^2$ gives the PDE
    $dS_t = rS_t dt + \sigma S_t dW_t$
-   Note that the drift parameter is now r, and the Wiener process
    $\bar W$ has distribution $\bQ$


# Extension of Arbitrage Theorem

1. **Savings Account**: the rate over $[t_1;t_2]$ is a constant $R_{t_1}$ and known at time $t_1$. The rate over the interval $[t_2;t_3]$, denoted by $R^u_{t_1}$ or $R^d_{t_2}$, is random and will not be known until $t_2$
2. **Forward Rate Agreement**: An FRA contract set at time $t_1$ where the payoff is determined by the rate at time $t_2$ and settled at $t_3$
3. **Short-Term Bond**: Short-term bond assumed to mature at time $t_3$
4. **Long Term Bond**: Long-term bond is assumed to mature at time T 
5. **Interest Rate Derivative**: The specific type of derivative is left general, and the payoff at $t_3$ will be denoted by $C_{t_3}$

Applying the same ideas as the early introduction to [arbitrage pricing](#FD), $\psi$ represents the discounted state probability and the below is only true if and only if there are no-arbitrage possibilities.


$$
\begin{bmatrix}
1 \\ 0 \\ B^s_{t_1} \\ B_{t_1} \\ C_{t_1}
\end{bmatrix} 
=
\begin{bmatrix}
R_{t_1}R^u_{t_2} & R_{t_1}R^u_{t_2} & R_{t_1}R^d_{t_2} & R_{t_1}R^d_{t_2} \\
F_{t_1}-L^u_{t_2} & F_{t_1}-L^u_{t_2} & F_{t_1}-L^du_{t_2} & F_{t_1}-L^d_{t_2} \\
1 & 1 & 1 & 1 \\
B^{uu}_{t_3} & B^{ud}_{t_3} & B^{du}_{t_3} & B^{dd}_{t_3} \\
C^{uu}_{t_3} & C^{ud}_{t_3} & C^{du}_{t_3} & C^{dd}_{t_3}
\end{bmatrix} 
\begin{bmatrix}
\psi^{uu} \\ \psi^{ud} \\ \psi^{du} \\ \psi^{du}
\end{bmatrix} 
$$


## Forward Measure

We can define under the Risk-Neutral Measure $\bQ$ the price of the arbitrage-free FRA rate:

$$
F_{t_1} = \frac{1}{E^{\bQ}\left[\frac{1}{(1+r_{t_1})(1+r_{t_2})}\right]}E^{\bQ} \frac{L_{t_2}}{(1+r_{t_1})(1+r_{t_2})}
$$

However doing so leads to some undersireable properties such as:

-   $F_t$ not being an unbiased estimator of $L_{t_2}$: $F_{t_1} \neq E^\bQ[L_{t_2}]$
-   The spot-rate terms inside the expectation take with respect to $\bQ$ cannot be factored out
-   The pricing formula for the FRA is nonlinear under $\bQ$

From the Revised Arbitrage Theorem, $B^s_{t_1} = \psi^{uu} + \psi^{ud} + \psi^{du} + \psi^{du}$. Repeating the same steps to that we did to get the risk neutral measure $\bQ$. 

$$
1 = \sum_{i,j}^{ud} \frac{\psi^{i,j}}{B^s_{t_1}} =  \sum_{i,j}^{ud} \pi^{ij}\rightarrow \pi^{ij} \\ \text{new forward measure: } \pi^{ij} \frac{\psi^{i,j}}{B^s_{t_1}}
$$

Under the new forward measure:

$$
F_t = E^\pi[L_T]
$$

## Other Measures

* Numeraire - a unit of commerce in which prices are measured
* 3 additional measures:
  + Share measure - numeraire is the underlying price $S_t$
  + Spot measure - numeraire is the "rolling over bank account" $B^*_t$
  + Swap measure - numeraire is the present value of a basis point (PVBP)

# Modeling Term Structure and Related Concepts

**Bond Yields and Bond Prices**

$B(t,T) = e^{-R(t,T)(T-t)}, t<T$

Solving for R(t,T) gives: $R(t,T) = \frac{-\log B(t,T)}{T-t}$

**Curve Definitions**

- Yield Curve: The spectrum of yields $R(t,T), T \in (t,T_{\max})$
- Discount Curve - The spectrum of default-free zero-coupon bond prices $B(t,T), t \in (t, T_{\max})$
- Credit Spread Curve: Given two yield curves:
  + $R(t,T), T \in (t,T_{\max})$ that correspond to default-free bonds (risk free bond yield)
  + $\bar{R}_t^T(t,T), T \in (t,T_{\max})$ that correspond to bonds with a given default probability (risky bond yield)
  + The spectrum of spreads, $s(t,T) = \bar{R}_t^T(t,T) - R(t,T) , T \in (t, T_{\max})$ is called the credit spread curve.
  
**Constant Spot Rates**

Assume that the spot rate $r_t$ is constant r, and also that we can lend and borrow at that rate. Then the price of a default-free zero-coupon bond paying \$1 at time T is: $B(t,T) = e^{-r(T-t)}$

**Bond Pricing Equation**

$$
B(t,T) = E^{\bQ}[e^{-\int_t^T r_s ds}] \\
R(t,T) = \frac{-\log B(t,T)}{T-t} = \frac{E^{\bQ}[e^{-\int_t^T r_s ds}]}{T-t} \\
F(t,T,U) = \frac{\log B(t,T) - \log B(t,U)}{U-T}
$$

- The bond price depends on the spectrum of future shor rates $t < s < T$
- The expectation is taken over a risk-neutral measure because the stochastic nature of interest rates add a risk premium for market risk

**Forward Rates**

Three times: t < T < U

Time t is when the rate is contracted. The cash transactions take place at times T and U. For example, if \$1 is loaned at time T, then the money returned at time U will be given by $e^{F(t,T,U)(U-T)}$

The bond pricing equation using the instantaneous forward rate is: $B(t,T) = e^{-\int_t^T F(t,s) ds}$

We can also write the forward rateds in terms of bond prices: $F(t,T,U) = \frac{\log B(t,T) - \log B(t,U)}{U-T}$ where as $T \rightarrow U$, the instantaneous forward rate is obtained: $F(t,T) = \lim_{T\rightarrow U} F(t,T,U)$

Two ways to accumulate from time T to time U: $e^{F(t,T,U)(U-T)} = \frac{B(t,T)}{B(t,U)}$

Taking the natural logs give: $F(t,T,U)(U-T) = \log B(t,T) - \log B(t,U)$

Rearranging yields: $F(t,T,U) = \frac{\log B(t,T) - \log B(t,U)}{U-T}$ 

## Classical vs HJM

**Markov Property**: states that the distribution of a process depends on the latest observation t, but not on the prior history at times s < t. 
- Markov, predicting stocks based on current price
- Non-Markov, momentum investing (requiring historical stock price)


**Classical Approach**: $B(t,T) = E^{\bQ}[e^{-\int_t^T r_s ds}]$

- Find a spot rate process $r_t$ that calibrates to observed market bond prices $B(t,T)$
- Two main ways to exploit the equation:
  + Use the bond prices $B(t,T)$ to obtain an arbitrage-free model for the spot rate $r_t$. Then one can exploit this model to price interest rate derivatives other than bonds
  + Can also posit a model about the spot rate $r_t$ from historical rates to get a set of prices for interest rate derivatives
- Suppose that we write down the following equation from $0 \leq i \leq n$
  + $B(t,T_i) = E^{\bQ}[e^{-\int_t^{T_i} r_s ds}]$
  + Use the n+1 bond prices to get a spot rate model $dr_t = a(r_t, t) dt + b(r_t, t) dV_t$
  
Geometric SDE Spot Rate Model

- $dr_t = \mu r_t dt + \sigma r_t dW_t$
- $a(r_t, t) = \mu r_t, b(r_t, t) = \sigma r_t$
- Disadvantages
  + If $\mu \neq 0$, then the model will diverge as $t \rightarrow \infty$
  + A constant volatility model may be oversimplified
  + Parameterization is difficult. There are two unknown parameters $\mu$ and $\sigma$ but n+1 equations.
  
Mean Reverting Spot Rate Model

- $dr_t = \lambda (\theta_t - r_t)dt + \sigma \sqrt{r_t} dW_t$
  + $\lambda$ is how quickly the mean reverts back
  + $\theta$ sets the level of mean reversion
- Since $\theta_t$ is a function of t, this increases the number of parameters. If $t_0 < t_1 < \cdots < t_m$, then there are a total of m+3 free parameters
- Advantages over Geometric SDE for interest-rate modeling
  + More free parameters which allow for more flexibilty
  + Under the right conditions, the mean-reverting model will not explode (with probability 1) as $t \rightarrow \infty$
  + The mean-reverting model above prevents negative interest rates given sufficiently small time steps.

**Heath-Jarrow-Morton Approach (HJM)**: $B(t,T) = e^{-\int_t^T F(t,s) ds}$

- HJM exploits the arbitrage relation between forward rates and bond prices to impose restrictions on the dynamics of the instantaneous forward rates directly.
- $F(t,s)$ is the instantaneous forward rate at time t of a loan from time s to  s + ds
- More general and less practical
- No explicit drift assumption needed
- Volatilities again will need to be calibrated
- Shows that in general the spot rate $r_t$ is not Markov

For HJ, we will get a forward rate model:

$$
dF(t,T) = a(F(t,T),T)dt + b(F(t,T), T)dW_t \\
\text{where:}  
\begin{cases}
a(F(t,T), T) = \sigma(t,T, B(t,T))\frac{\partial \sigma(t,T,B(t,T))}{\partial T} \\
b(F(t,T),T) = \frac{\partial \sigma(t,T,B(t,T))}{\partial T}
\end{cases}
$$

This approach derives an exact no-arbitrage restriction on the drift coefficient for instantaneous forward rate dynamics. This is similar to Black-Scholes, where the drift is replaced with a more complicated term that depends on the volatilities of the bonds under consideration.

HJM Equations for dB and r: to be used to get bond price process and spot rate model r.

$$
dB(t,T) = r_t B(t,T) dt + \sigma(t,T, B(t,T)) B(t,T) dW^{\bQ}_t \\
r_t = F(0,t) + \int^t_0 b(s,t) [\int^t_s b(s,u)du]ds + \int^t_0 b(s,t)dW_s 
$$

We can see from the equation that $r_t$ does not depend on prior history at times s < t[^2]. Because $r_t$ is a complex function of all past forward rate volatilities making in non-Markovian. Current empirical work indicates that spot-rate behavior in reality may fail to be Markovian, so this gives flexibility to market practitioners. Being Non-Markovian is an advantage of HJM.

[^2]: the s in $b(s,u)$ represents $F(t,T)$ which is a prior forward rate. 

**HJM vs Classical**

- Classical approach is used more in practice.
- HJM may have explosiveness problems for finite time, which is very undesirable.
- Much more familiarity with spot rate models, so this partially explains the popularity of classical approach.

## Fitting to Initial Term Structure <span style="color: red;">[FLAG] Review this section again, or combine because it's too short? <span style="color: red;">

1. Monte Carlo
2. Tree Models
3. Closed-Form Solutions

## Classical PDE Analysis for Interest Rate Derivatives

Consider two bonds with prices denoted $B(t,T_1), B(t,T_2)$ driven by the same Wiener process where $T_1 < T_2$. Suppose that bond prices are postulatd to behave according to he following processes:

$$
dB^1 = \mu_1(B^1,t)B^1dt + \sigma_1(B^1,t)B^1dW_t \\
dB^2 = \mu_2(B^2,t)B^2dt + \sigma_2(B^2,t)B^2dW_t
$$

Notice that both bonds are influenced by the same Wiener process, $W_t$. Additionaly, suppose that the interest rate model is given by:

$$
dr_t = a(r_t, t) dt + b(r_t, t)dW_t
$$

Step 1. Create a portfolio of two bonds:

$$
P = \theta_1 B^1 - \theta_2 B^2 \\
dP = \theta_1 dB^1 - \theta_2 dB^2 \\ 
$$

Suppose the portfolio positions are chosen as $\theta_1 = \frac{\sigma_2}{\sigma_2 - \sigma_1}\frac{P}{B^1}, \theta_2 = \frac{\sigma_1}{\sigma_2 - \sigma_1}\frac{P}{B^2}$. Plugging into the equation for $dB^1$ and $dB^2$ to get:

$$
dP = \theta_1 [\mu_1 B^1 dt + \sigma_1 B^2 dW_t] - \theta_2 [\mu_2 B^2 dt + \sigma_2 B^2dW_t]
$$

Grouping the drift and innovation terms gives:

$$
dP = [\theta_1 \mu_1 B^1 - \theta_2 \mu_2 B^2]dt + [\theta_1 \sigma_1 B^1 - \theta_2 \sigma_2 B^1]dW_t
$$

Given $\theta_1$ and $\theta_2$, we have that $\theta_1 \sigma_1 B^1 - \theta_2 \sigma_2 B^2 =0$. Thus, the change in portfolio value is given by:

$$
dP = [\theta_1 \mu_1 B^1 - \theta_2 \mu_2 B^2]dt = \frac{\sigma_2 \mu_1 - \sigma_1 \mu_2}{\sigma_2 - \sigma_1} Pdt = r_t Pdt
$$

The last equality follows by the standard no-arbitrage argument, since we have formed a risk-free portfolio with no innovation term.

$$
\frac{\sigma_2 \mu_1 - \sigma_1 \mu_2}{\sigma_2 - \sigma_1} = r_t \rightarrow \frac{\mu_1 - r_t}{\sigma_1} = \frac{\mu_2 - r_t}{\sigma_2} \\
\text{Market Price of Interest Rate Risk:} \frac{\mu_i - r_t}{\sigma_i} = \lambda(r_t,t)
$$

- The Sharpe ratio of the bonds, given by the Market Price of Interest Rate risk, are equal
- Risk premia of per unit volatility are the same across bonds
- Keep in mind, this is based on the assumption that bonds are subject to the same source of risk $W_t$
- The market price of interest rate risk equation holds regardless of the time to maturity of the bond
- We can create a portfolio of two bonds with weights set such that the portfolio is riskless. By a no-arbitrage argument, we get that the portfolio must earn the risk-free rate. This led to the conclusion that the Sharpe ratio of all bonds must be equal in a no-arbitrage framework

### Derivation of Bond Price PDE

Recall the interest rate model:

$$
dr_t = a(r_t, t)dt + b(r_t,t)dW_t
$$
Our expression for $dB(t,T)$:

$$
dB(t,T) = B_r dr_t + B_t dt + \frac{1}{2}B_{rr}(dr_t)^2 \\
dB(t,T) = B_r[a(r_t,t)dt + b(r_t,t)dW_t] + B_tdt + \frac{1}{2}B_{rr}b(r_t,T)^2dt \\
dB(t,T) = [B_ra(r_t,t) + B_t + \frac{1}{2}B_{rr}b(r_t,T)^2]dt + B_r b(r_t,t)dW_t
$$

For simplicity, we write the SDE as:

$$
dB(t,T) = \mu (B,t) Bdt + \sigma(B,t_ B)dW_t \text{, where:}
\begin{cases}
\sigma(B,t)B = B_r b(r_t,t) \\
\mu (B,t) B = B_ra(r_t,t) + B_t +\frac{1}{2}B_{rr}b(r_t,T)^2
\end{cases}
$$

Using the market price of interest rate risk equation gives that:

$$
\frac{\mu (B,t) - r_t}{\sigma(B,t)} = \frac{\mu (B,t)B - r_tB}{\sigma(B,t)B} = \lambda_t \rightarrow \mu(B,t)B - r_t B = \lambda_t \sigma(B,t)B
$$

Plugging in for $\mu$ and $\sigma$ equations and rearranging gives us the key bond PDE. It has a simple boundary condition given by $B(T,T) = 1$

$$
B_r[a(r_t,t) - \lambda_tb(r_t,t)] + B_t + \frac{1}{2}B_{rr}b(r_t,T)^2 - r_tB = 0
$$

### Comparing to Black-Scholes

| Model      |  Randomness   | PDE |  Boundary Condition  |
|:----------:|:----------:|:----------:|:----------:|
| Trad BS Model | $S_t$ |   $\frac{\partial F}{\partial t} + \frac{\sigma^2 S^2}{2}\frac{\partial^2F}{\partial S^2} + rS \frac{\partial F}{\partial S} - rF= 0$  |        $F(S_T,T) = \max[S_T - K,0]$        | 
|  Bond Pricing  |  $r_t$|   $B_r[a(r_t,t) - \lambda_tb(r_t,t)] + B_t + \frac{1}{2}B_{rr}b(r_t,T)^2 - r_tB = 0$   | $B(T,T) = 1$ |

**Trad BS Model**

- Assumes that the instantaneous risk-free spot rate $r_t$ is constant
- The randomness is driven by the stock price process $S_t$
- No need to explicitly model the drift term of the stock price process $S_t$
- The risk-free portfolio is created by combining lending/borrowing, an options position, and shorting/longing the underlying instrument
- Option price depended on relevant volatilities only

**Bond Price PDE**

- Does not assume that the instantaneous risk-free spot rate $r_t$ is constant
- Randomness is driven by the spot rate process $r_t$
- Thus, the spot rate process $r_t$ is involved in the bond pricing PDE
- The risk-free portfolio is created by adjusting the weights in two bonds
- Calibration and estimation is more challenging here, because the bond pricing PDE is more complex needing to estimate two new parameters:
  + the spot rate drift $a(r_t,t)$
  + the market price of interest rate risk $\lambda_t$
- Made the simplifying assumption that all bonds are driven by the same Wiener Process $W_t$. This led us to having a single market price of risk $\lambda_t$

# Term Structure Models - Continuous Time

Introduction to stochastic interest rate models: **Vasicek**

$$
dr_t = \gamma(\bar{r} - r_t) dt + \sigma dX_t
$$

- $\bar{r}$ is the long term value of the short rate
- $\gamma$ is the speed of the mean reversion
- $\sigma$ is the volatility of the short rate r(t)

Solving for $r_t$, conditional at a time s<t:

$$
r(t) = r(s) e ^{\gamma(t-s)} + \bar{r} (1 - e ^{\gamma(t-s)}) + \sigma \int_s^t e ^{\gamma(t-u)} dX(u)
$$

The stochastic integral is normally distributed thus r(t) is normally distributed with:

$$
\text{Expectation: } E[r_t|F_s] = \bar{r} + (r_0 - \bar{r})e ^{-\gamma(t-s)} \\
\text{Variance: } Var[r_t|F_s] = \frac{\sigma ^2}{2\gamma}[1 - e ^{-2\gamma(t-s)}]
$$

As $t \rightarrow \infty$, the stationary distribution of $r_t$ is still normal with:

- Mean: $\bar{r}$
- Variance: $\frac{\sigma ^2}{2\gamma}$

## No-Arbitrage Pricing: Vasicek Model

1. Setup a portfolio that is long on eunit of a security with price $Z_1(r,t)$ and short $\Delta$ units of another interest rate security with price $Z_2(r,t)$
    - The value of this portfolio at time t: $\Pi(r,t) = Z_1(r,t) - \Delta Z_2(r,t)$
2. Choose $\Delta$ so that interest rate risk is hedged
    - $\frac{\partial \Pi(r,t)}{\partial r} = \frac{\partial Z_1(r,t)}{\partial r} - \frac{\partial \Delta Z_2(r,t)}{\partial r}$
    - $\Delta = \frac{\partial Z_1 / \partial r}{\partial Z_2 / \partial r}$
3. Apply Ito's Lemma to solve for the stochastic dynamics of $\Pi(r,t)$

$$
\begin{aligned}
d\Pi_t &= \frac{\partial \Pi}{\partial t} dt +\frac{\partial \Pi}{\partial r} dr+ \frac{1}{2}\frac{ \partial^2 \Pi}{\partial r^2}dr^2 \\
&= \left(\frac{\partial \Pi}{\partial t} + \frac{\partial \Pi}{\partial r}\gamma(\bar{r} - r_t) + \frac{1}{2}\frac{ \partial^2 \Pi}{\partial r^2}\sigma^2\right)dt + \frac{\partial \Pi}{\partial t}\sigma dX_t
\end{aligned}
$$

Under the hedging strategy, $\frac{\partial \Pi}{\partial r} = 0$, so $d\Pi$ simplifies to:

$$
d\Pi_t = \left(\frac{\partial \Pi}{\partial t} + \frac{1}{2}\frac{ \partial^2 \Pi}{\partial r^2}\sigma^2\right)dt
$$

4. Impose a no-arbitrage constraint based on the fact that the portfolio must earn the risk-free rate: $d\Pi_t = r_t \Pi dt$ so that:

$$
d\Pi_t = \left(\frac{\partial \Pi}{\partial t} + \frac{1}{2}\frac{ \partial^2 \Pi}{\partial r^2}\sigma^2\right)dt = r_t \Pi_t 
$$

5. Rewrite the no-arbitrage conditions in terms of the two securities $Z_{1,t}$ and $Z_{2,t}$ and utilize the fact that $\Delta = \frac{\partial Z_1 / \partial r}{\partial Z_2 / \partial r}$ : 

$$
\frac{\partial Z_1}{\partial t} + \frac{1}{2} \frac{\partial^2 Z_1}{\partial r^2}\sigma^2 - r_tZ_1 = \Delta \left(\frac{\partial Z_2}{\partial t} + \frac{1}{2} \frac{\partial^2 Z_2}{\partial r^2}\sigma^2 - r_tZ_2\right) \\
\frac{\frac{\partial Z_1}{\partial t} + \frac{1}{2} \frac{\partial^2 Z_1}{\partial r^2}\sigma^2 - r_tZ_1}{\partial Z_1 / \partial r} = \frac{\left(\frac{\partial Z_2}{\partial t} + \frac{1}{2} \frac{\partial^2 Z_2}{\partial r^2}\sigma^2 - r_tZ_2\right)}{\partial Z_2 / \partial r}
$$

6. $-m^*(r,t) = \frac{\frac{\partial Z}{\partial t} + \frac{1}{2} \frac{\partial^2 Z}{\partial r^2}\sigma^2 - r_tZ}{\partial Z / \partial r}$
  - The Vasicek model assumes $m^*(r,t)$ has the form $m^*(r,t) = \gamma^*(\bar{r}^* - r)$
7. Rearrange to obtain the **fundamental pricing equation** that all interest rate derivatives must satisfy:

$$
\frac{\partial Z}{\partial t} + \frac{\partial Z}{\partial r} m^*(r,t) + \frac{1}{2} \frac{\partial^2 Z}{\partial r^2}\sigma^2 = rZ
$$

- This partial differential equation was derived from a no-arbitrage condition
- Solving this equation, subject to a final condition for the value of the security at T, will yield a pricing formula for any interest rate security $Z(r,t;T)$.
- The final condition of the value of the security at maturity T is what differentiates one interest rate security from another.
  + Zero-Coupon Bond: $Z(r,t;T) = 1$
  + Interest rate option at strike $r_K$: $Z(r,t;T) = N \times \max{(r-r_K,0)}$

### Vasicek Zero Coupon Bond

Solution to the fundamental pricing equation under a Vasicek model for a zero-coupon bond is:

$$
Z(r,t;T) = e^{A(t;T) - B(t,T) \times r} \\
A(t;T) = (B(t;T) - (T-t)) \left(\bar{r}^* - \frac{\sigma ^ 2}{2(\gamma^*)^2}\right) - \frac{\sigma^2 B(t;T)^2}{4\gamma^*} \\
B(t;T) = \frac{1}{\gamma^*}(1-e^{-\gamma^* (T-t)})
$$

Under the Vasicek model, the spot rate, $r_t(\tau)$ for maturity $\tau$ is:

$$
r_t(\tau) = -\frac{\ln(Z(r_t, \tau))}{\tau} = - \frac{A(\tau)}{\tau} + \frac{B(\tau)}{\tau}r_T
$$

 - When $r_t$ is low, the Vasicek model implies a rising term structure, likewise, when $r_t$ is high, the model implies decreasing term structure
 - As $r_t$ changes, all the points on the term structure will move with perfect correlation which is true for all one-factor interest rate models in general, and is considered a large drawback of one factor models.
 
**Applying Ito's Lemma on $r_t(\tau)$:**

$$
dr_t(\tau) = \left(\frac{B(\tau)}{\tau} \gamma(\bar{r} - r_t)\right)dt + \frac{B(\tau)}{\tau} \sigma dX_t
$$

Note that $\frac{B(\tau)}{\tau}$ just acts as a scalar for the mean reversion speed and diffusion coefficient. For a fixed value of $\gamma^*$, this scalar decreases as maturity increases.

- $\frac{B(\tau)}{\tau}$ is in the diffusion coefficient. Longer-term rates are **less** volatile than shorter term rates.
- $\frac{B(\tau)}{\tau}$ is the mean reversion speed. Longer-term rates move **slower** than short-ter rates in converging to the long-term rate $\bar{r}$
- For a given time-to-maturity $\tau$, $B(\tau)$ decreases as $\gamma^*$ increases.
- Ultimately Lower levels of $\gamma^* \rightarrow$ higher $B(\tau) \rightarrow$ higher variation in the long-term spot rates.

### Spot Rate Duration

$$
D_Z = -\frac{1}{Z} \frac{\partial Z}{\partial r}
$$

Under the Vasicek model, $\frac{\partial Z}{\partial r}$ is given by:

$$
\frac{\partial Z}{\partial r} = -B(\tau) \times Z
$$

The spot rate duration of a zero-coupon bond with time to maturity $\tau$ is given by:

$$
D_Z(\tau) = B(\tau) = \frac{1}{\gamma^*}(1-e^{-\gamma^* \tau})
$$

Lower $\gamma^*$ = higher spot-rate duration. Remember that $\gamma^*$ is the risk neutral parameter of $\gamma$, not the real world value of $\gamma$.

### Coupon Bond

Since a coupon bond is a weighted average of zero-coupon bond bonds, it follows that the spot rate duration of a coupon bond under the Vasicek model is:

$$
D_P = \sum_{i=1}^n w_i D_Z (\tau_i)
$$

The weights for $w_i$ are given by for $i = 1, \dots,n$ : $w_i = 100 \frac{c}{2} \frac{Z(r,\tau_i)}{P(r,\tau)}$

The weight for $i=n$, denoted by $w_n$ is: $w_n = 100 \left(1+\frac{c}{2}\right)\frac{Z(r,\tau_n)}{P(r,\tau)}$

### Estimating Parameters of Vasicek Model

$\sigma, \bar{r}, \gamma$ can be estimated from historical data.

 - $\sigma$: calculate the historical volatility of the short rate $r_t$
 - $\bar{r}$: average the short-term rate over the sample historical period
 - $\gamma$: the slope from regressing the changes in the interest rate $(r_{t+\delta} - r_t)$ on $r_t \times \delta$
  + If we ignore the stochastic component of the Vasicek model, the change in $r_t$ is: $dr_t = \gamma \times \bar{r} - \gamma \times r_t \times dt$
 - $\bar{r}^* and \gamma^*$ are calibrated to a set of zero-coupon bonds observed in the market. We solve for the set $(\bar{r}^*, \gamma^*)$ that minimizes the following:
 
 $$
 J(\bar{r}^*, \gamma^*) = \sum_{i=1}^n (Z^{\text{Vasicek}}(r_0, 0, \bar{r}^* , \gamma^*; T_i) - Z^\text{Data}(0, T_i))^2
 $$
- $Z^\text{Data}(0, T_i)$: the market price for a zero-coupon bond price expiring at $T_i$
- $Z^{\text{Vasicek}}(r_0, 0, \bar{r}^* , \gamma^*; T_i)$: the Vasicek model's zero-coupon bond price.
- Formula essentially is minimizing the absolute difference between real world and Vasicek model by guessing $\bar{r}^*, \gamma^*$

### Pricing Options on Zero-Coupon Bonds

An option on a zero-coupon bond with maturity $T_O$ and strike price K has a payoff of:

$$
\text{Payoff of } T_O = max(Z(T_O; T_B) - K, 0)
$$

- $Z(T_O; T_B)$ is the value at time $T_O$ of the bond with maturity $T_B$ under the Vasicek model, the call option value is:

$$
V(r_0,0) = Z(0, r_0;T_B)N(d_1) - KZ(0,r_0; T_O)N(d_2)
$$

- $N(x)$ is the cumulative standard normal distribution of $d_1$ and $d_2$ given by:

$$
d_1 = \frac{1}{S_Z(T_O)} \log \left(\frac{Z(0, r_0;T_B)}{KZ(0,r_0; T_O)}\right) + \frac{S_Z(T_O)}{2} \\
d_2 = d_1 - S_Z(T_O) \\
S_Z(T_O) = B(T_O; T_B)\sqrt{\frac{\sigma^2}{2\gamma^*}(1 - e^{-2 \gamma^* T_O})}
$$

- $S_Z(T_O)$ is the standard deviation of the log return between 0 and $T_O$ of the underlying zero coupon bond.
- Usin gthe fact that $Var(r(T_O)) = \frac{\sigma^2}{2\gamma^*}(1 - e^{-2 \gamma^* T_O})$ under the Vasicek model, we an solve for $S_Z(T_O)$ as:

$$
\begin{aligned}
Var[\log \left(\frac{Z(T_O, r_{T_O};T_B)}{Z(0,r_0; T_B)}\right)] &= Var[log(Z(T_O, r_{T_O}; T_B))] \\
 &= Var[A(T_O;T_B) - B(T_O;T_B)r_{T_O}] \\
 &= B(T_O; T_B)^2Var[r_{T_O}] \\
 &= B(T_O;T_B)^2 \frac{\sigma^2}{2\gamma^*} (1-e^{-2 \gamma^* T_O})
\end{aligned}
$$

### Problems

 - The model parameters need to be calibrated to market zero-coupon bond prices.
    + Even after calibration is complete $Z^\text{Vasicek}(0,T) \neq Z^\text{Mkt}(0,T)$
 - Example, Pricing option on zero-coupon bond. We run into issues
    + Do we use market discount factors, or model discount factors?
 - Solution: Use no-arbitrage models by having the market and model zer-coupon bond prices be the same
    + [Ho-Lee Model](#HoLee)
    + [Hull-White MOdel](#HullWhite)

## CIR Model

$$
dr_t = \gamma(\bar{r} - r_t) dt + \sqrt{\alpha r_t} dX_t
$$

- Drawback of Vasicek Model; Interest rates can be negative
- Under the CIR model, the interest rate is positive as long as $\gamma \bar{r} > \frac{1}{2} \alpha$
  + When $r_t$ hits 0, the diffusion coefficient becomes 0
  + $\gamma \bar{r}$ must be large enough so that it can pull up the interest rate back to the long term rate $\bar{r}$
- The distribution of $r_t$ is a noncentral chi-square distribution
- One can also derive analytical formulas for interest rate securities under the CIR model

## Replicating Portfolio and Relative Value Trades

The return of security 1 can be replicated by holding $\Delta$ units of security 2, Investing in a cash position $C_t = Z_1(r_t,t) - \Delta Z_2(r_t,t)$ at the risk free rate. Where the value of the replicating portfolio for $Z_1$ at time t is:

$$
P_t = \Delta Z_{2,t} + C_t
$$

By construction, $P_t$ and $Z_{1,t}$ have the same value.

The value of $\Delta = \frac{\partial Z_1 / \partial r}{ \partial Z_2 / \partial r}$ at the next timestep t + dt will change. To rebalance, adjust the portfolio position in security 2 to the new $\Delta_{t+dt}$. The new position in cash in our replicating portfolio at t+dt is:

$$
C_{t+dt} = C_t + C_t r_t dt - (\Delta_{t + dt} - \Delta_t) Z_{2,t + dt}
$$

After rebalancing the new value of the replicating portfolio is.

$$
P_{t+dt} = \Delta_{t+dt} Z_{2,t+dt} + C_{t+dt}
$$

One can setup a **relative value trade** to exploit a potential arbitrage opportunity when prices of interest rates securities don't align with prices predicted by an interest rate model. To setup a relative value trade:

1. Sell short a STRIP
2. Construct a replicating portfolio of the strip using the hedge ratio from the Vasicek model.
    -   Go long $\Delta$ units of the strip with a longer maturity.
    -   Invest a cash position at the risk free rate.
    -   Generally you want to pick the longer maturity at a time where the Vasicek model had low pricing error.
3. Dynamically rebalance this trading strategy overtime.
4. Assuming our estimated model is correct, we should be perfectly tracking the value of the initial STRIP that we are short.

## Risk Premiums

Start with Vasicek Model Dynamics

$$
dr_t = \gamma(\bar{r} - r_t) dt + \sigma dX_t
$$

Let Z(r,t) be teh price at time t of a zero-coupon bond and apply Ito's Lemma, the dynamics of Z(r,t) are given by:

$$
dZ = [\frac{\partial Z}{\partial t} + \frac{1}{2}\frac{\partial ^2 Z}{\partial r^2} \gamma (\bar{r} - r)]dt + \frac{\partial Z}{r} \sigma dX_t \\
E[dZ] = [\frac{\partial Z}{\partial t} + \frac{1}{2}\frac{\partial ^2 Z}{\partial r^2} \gamma (\bar{r} - r)]dt
$$

from the Fundamental Pricing Equation:

$$
\frac{\partial Z}{\partial t} + \frac{1}{2}\frac{\partial ^2 Z}{\partial r^2} \sigma^2 = rZ - \frac{\partial Z}{\partial r} \gamma^* (\bar{r}^* - r)
$$

Substitute in for $\frac{\partial Z}{\partial t} + \frac{1}{2}\frac{\partial ^2 Z}{\partial r^2} \sigma^2$ and utilize the fact that $\frac{\partial Z}{\partial r} = -B(r,T)Z$

$$
E[dZ] = rZdt - B(t;T) Z(\gamma(\bar{r} - r) - \gamma^* (\bar{r}^* -r))dt
$$

Rearranging terms, we obtain the annualized **risk premium formula** for a zero-coupon bond under the Vasicek model (aka expected excess return):

$$
\text{Risk Premium} = E[\frac{dZ}{Z}]/dt - r = -B(t;T)(\gamma(\bar{r} - r) - \gamma^* (\bar{r}^* -r))dt
$$

**Market Price of interest rate risk**

$$
\lambda(r,t) = \frac{1}{\sigma} (\gamma(\bar{r} - r) - \gamma^* (\bar{r}^* -r)) \\
\lambda_0 = \frac{1}{\sigma} (\gamma\bar{r} - \gamma^*\bar{r}^*) , \lambda_1 = \frac{1}{\sigma} (\gamma^* - \gamma) \\
\text{rewrite: } \lambda(r,t) = \lambda_0 + \lambda_1 r
$$

The stochastic dynamics of dZ/Z can be written as $\frac{dZ}{Z} = \mu_Z dt + \sigma_Z dX$ where:

$$
\sigma_Z = \frac{1}{Z} \frac{\partial Z}{\partial r} \sigma = -B(t;T)\sigma \\
\mu_Z = \frac{1}{Z}[\frac{\partial Z}{\partial t} + \frac{1}{2} \frac{\partial^2 Z}{ \partial r^2}\sigma^2 + \frac{\partial Z}{ \partial r} \gamma (\bar{r} - r)]
$$

By multiplying the numerator and denominator of the risk premium by $\sigma$, then plugging in $\sigma_Z$ we can see that the risk premium is:

$$
\text{Risk Premium} = \sigma_Z \times \lambda(r,t)
$$

**Geeneral Interest Rate Model**

Consider a general interest rate model with dynamics:

$$
\begin{matrix}
\text{General Interest Rate Model} & dr_t = m(r_t,t)dt + s(r_t,t)dX_t \\
\text{Risk premium on: }Z(t,r) & E[\frac{dZ}{Z}]/dt - r = \sigma_Z \times \lambda(r,t) \\
\text{Volatility of the log-return onf the security} & \sigma_Z = \frac{1}{Z}\frac{\partial Z}{\partial r} s(r,t) \\
\text{Market price of risk:} & \lambda(r,t) = \frac{1}{s(r,t)}(m(r,t)-m^*(r,t))
\end{matrix}
$$

Risk Neutral Dynamics are primarily used for pricing interest rate securities and are not realistic scenarios.

Risk Natural Dynamics, AKA real world scenarios are realistic. It uses true risk parameters and is often used for risk analysis.

# Macroeconomic Model of Term Structure

| Values     |  equation  | dynamics |
|:----------:|:----------:|:----------:|
| GDP | $y_t = ln(Y_t)$ |   $dy_t = gdt + \sigma_y dX_{y,t}$  |
| CPI  |  $q_t = ln(Q_t)$ |   $dq_t = i_t dt + \sigma_q dX_{q,t}$  |
| expected inflation | $i_t$ | $di_t = \gamma(\bar{i} - i_t)dt + \sigma_i dX_{i,t}$ |

Correlations are: $\rho_{qy} = dX_{y,t}dX_{q,t} ;\rho_{iq} = dX_{i,t}dX_{q,t}; \text{and} \rho_{yi} = dX_{y,t}dX_{i,t}$

Utility Function: The utility a market participant receives from consuming $C_t$ units of the CPI index. $\rho$ is a parameter that reflects people's behavior to consume earlier than later. h is a parameter that describes a persons risk averision. higher h = more risk averse, 0 h is a risk neutral person.

$$
U(C_t,t)=e^{-\rho t}\frac{C_t^{1-h}}{1-h}
$$

The market participant has two choices to spend $1.

1. Spend \$1 today and buy $\frac{1}{Q_t}$ units of CPI.
    -   This increases the consumer utility by $\frac{\partial U(C_t,t)}{\partial C Q_t}$
2. Invest \$1 to buy $\frac{1}{Z(t,T)}$ units of zero coupon bond.
    -   Allows the consumer to buy $\frac{1}{Z(t,T) Q_T}$ of CPI, increasing utility at the future time T to $\frac{1}{Z(t,T)}\frac{\partial U(C_T,T)}{\partial C_T Q_T}$
3. This leads to an equilibrium condition: $\frac{\partial U(C_t,t)}{\partial C Q_t} = \frac{1}{Z(t,T)}\frac{\partial U(C_T,T)}{\partial C_T Q_T}$
4. Solving for what the theoretical zero coupon bond price should be: $Z(t,T) = E_t [\frac{Q_t \partial( U(C_T,T) / \partial C_T}{Q_T \partial U(C_t, t) / \partial C_t}]$
5. Plug in the utility function directly to get: $Z(t,T) = E_t [\frac{Q_t C_T^{-h} e^{-\rho T}}{Q_T C_t^{-h} e^{-\rho t}}]$
6. Assume that $C_t = kY_t$ and plug in the dynamics for Q and Y to then get:

$$
Z(i,t,T) = e^{A(t;T) - B(t;T)(i+c)} 
\begin{cases}
B(t;T) = \frac{1}{\gamma}(1 - e^{-\gamma(T-t)}) \\
A(t;T) = \left(B(t;T) - (T-t))(\bar{r}^* - \frac{\sigma_i^2}{2\gamma^2}\right) - \frac{\sigma^2_i B(t;T)^2}{4\gamma}
\end{cases}
$$

The spot nominal rate is given by the sum of expected inflation, $i_t = E[dq]$, and a constant c: $r_t = i_t + c$

**Additional formulas:**

$$
\bar{r} = \bar{i} + c \\
c = \underbrace{(\rho + hg - \frac{1}{2} h^2 \sigma^2_y)}_\text{measure of the real rate of interest} - \underbrace{h \sigma_y \sigma_q \rho_{qy}}_\text{inflation risk premium} - \frac{1}{2} \sigma^2_q \\
\bar{r}^* = \bar{r} - \frac{1}{\gamma}(h\sigma_i\sigma_y \rho_{yi} + \sigma_i \sigma_y \rho_{iq}) \\
\text{Dynamics of the spot nominal rate } r_t: \\
\text{Risk natural:} dr_t = \gamma(\bar{r} - r_t) dt + \sigma_i dX_i \\
\text{Risk neutral:} dr_t = \gamma(\bar{r}^* - r_t)dt + \sigma_i dX_i \\
\text{Market price of risk:} \lambda = \frac{\gamma}{\sigma_i}(\bar{r} - \bar{r}^* = h\sigma_y\rho_{yi} + \sigma_y)
$$

 - $\rho$: Higher means people earn more utility by consuming today. This results in more borrowing which increases the real rate of interest
 - hg: When the growth rate of the economy g is high, then people want to borrow to consume today, thus increasing the real interest rate.
 - $\frac{1}{2}h^2 \sigma^2_y$: When volatility of GDP growth is high, there is higher risk there will be less consumption in the future.
 - $\sigma_y \sigma_q \rho_{qy}$: Covariance of GDP growth (dy) and inflation (dq). When covariance is negative, a short term nominal investment rate $r_t$ will be more risky. Thus the nominal rate must increase to compensate for this risk.
 - $-h\sigma_i\sigma_y \rho_{yi}$: is an adjustment to the long-term unconditional average nominal rate $\bar{r}$
 - Since covariance ($\sigma_i\sigma_y \rho_{yi}$) between inflation and GDP is typically negative, this implies that the adjustment for risk is normally positive
 - $\bar{r}^*$ typically increases when:
    + The risk aversion h increases
    + The covariance cov(di,dy) becomes more negative
    + The mean reversion speed $\gamma$ decreases (is slower)

# Ho-Lee Model {#HoLee}

$$
dr_t = \theta_t dt + \sigma dX_t
$$

 - The parameter $\theta_t$ is a deterministic *function* that is chosen to match the market zero-coupon bond term structure.
 - $r_t$ is normally distributed
 - No mean-reversion, so future distribution of $r_t$ is unbounded
 - Rates can be negative
 
$$
Z(r,0; T) = e^{A(0,T)-T \times r} \\
A(0,T) = -\int_0^T (T-t) \theta_T dt + \frac{T^3}{6} \sigma^2 \\
\theta_t = \frac{\partial f(0,t)}{\partial t} + \sigma^2 \times t \\
\text{}
$$

Substituting in $\theta_t$ we can solve for A(0,T) analytically:

$$
A(t;T) = \ln \left( \frac{Z(r_0, 0;T)}{Z(r_0, 0;t)}\right) + (T-t)f(0,t) -\frac{\sigma^2}{2}(T-t)^2t
$$

## Approximating $\theta_t$

 1. Given the market STRIP zero-coupon bond prices, derive corresponding spot rates for each tenor T as: 
 
 $$
 r(0,T) = -\frac{\ln(Z^\text{Mkt}(0,T))}{T}
 $$

 2. Interpolate the spot rates in order to obtain a twice differentiable spot rate curve
 3. Compute the instantaneou forward curve numerically as:
 
 $$
 f(0,t) \approx r(0,t) + t \times \frac{r(0, t+ \delta) - r(0,t)}{\delta}
 $$

4. Approximate the deterministic function $\theta_t$ as:

$$
\theta_t \approx \frac{f(0, t+ \delta) - f(0,t)}{\delta} + \sigma^2 \times t
$$

## Term Structure of Volaitility

 - The yield $r_t(\tau)$ on a zero-coupon bond with maturity $\tau$ is: $r_t(\tau) = -\ln (Z(r,t;t + \tau))/\tau$
 - Plugging in the formula for $Z(r,t; t+\tau)$, $r_t(\tau) = -A(t;t+\tau)/\tau + r_t$
 - Taking the differentials of both sides results in the annualized volatility, $dr_t(\tau)$, denoted as $\sigma_t (\tau) = stdev(dr_t(\tau)) / \sqrt{dt} = \sigma$
 - The Ho-Lee model implies a flat term structure of volatility of change in spot rates. However, in the market, the volatility term structure is typically downward sloping
 
 
## Summary

**Advantages**

- Simple and has closed form solutions for many interest rate derivatives.
- A no-arbitrage model that is able to reprice market-zero coupon bond prices exactly

**Disadvantages**

- Non constant short rate $r_t$, as in the interest rates can grow to $\pm \infty$
- Implies a flat term structure of volatility of changes in bond yields, which is inconsistent with what is observed in the market.
- One factor model; assumes all points on the yield curve are perfectly correlated with each other.


# Theory

## Doob-Meyer Theorem

Suppose that $X_t , 0 \leq t \leq \infty$, is a right-continuous
submartingale w.r.t the family $\lbrace I_t \rbrace$ and
$E(X_t) < \infty$ for all t. Then $X_t$ admits the decomposition
$X_t = \mu_t + A_t$ where:

-   $\mu_t$ is a right-continuous martingale w.r.t P
-   $A_t$ is an increasing process measurable w.r.t $I_t$

## Riemann-Stieltjes

A generalization of the Riemann integral that can be denoted as:

$$
\int_{x=a}^b f(x)dg(x)
$$

Using a sequence of partitions $P = {a = x_0 < x_1 < \dots < x_n = b}$
we can approximate the integral to be the sum:

$$
S(P,f,g) = \sum^{n-1}_{i=0} f(c_i)[g(x_{i+1}) - g(x_i)]
$$

## Girsanov Theorem {#Girsanov}

### Preliminary Setup

Define a random process and impose a cndontion in $X_t$ where it cannot
vary "too much"

$$
\begin{matrix}
\xi_t = e^{\int_0^tX_udW_u - \frac{1}{2} \int_0^t X^2_udu} & t \in [0,\infty) \\
E[e^{\int_0^t X^2_udu}] < \infty  & t \in [0,\infty)
\end{matrix}
$$

Using Ito's Lemma, we can calculate

$$
d\xi_t = [e^{\int_0^tX_udW_u - \frac{1}{2} \int_0^t X^2_udu}][X_tdW_t] \rightarrow d\xi_t = \xi_tX_t dW_t
$$

By setting t = 0 we see $\xi_0 = 1$. Thus by taking the stochastic
integral of the above becomes:

$$
\xi_t = 1 + \int_0^t \xi_sX_sdW_s
$$

$\int_0^t \xi_sX_sdW_s$ is a stochastic integral with respect to a
Wiener process and is $I_t$ adapted and does not move rapidly. Implying
that this integral is a square integrable martingale.

### Theorem

If the process $\xi_t$ defined by
$\xi_t = e^{(\int_0^t X_u dW_u - \frac{1}{2}\int_0^t X^2_u du)}$ is a
martingale with respect to information sets $I_t$, and the probability
$\bP$, then $W^{*}_t$ , defined by

$$
W^{*}_t = W_t - \int_0^t X_u dW_u \quad t \in [0, \infty)
$$

is a Wiener process with respect to $I_t$ and with respect to the
probability measure $\bQ$ given by

$$
\bQ(A) = E^\bP(1_A\xi_T)
$$

with A being an event determined by $I_T$ and $1_A$ being the indicator
function of the event.

All this to say that basically multiplying a Wiener process with a
probability distribution $\xi_t$ gives us a new Wiener process. The two
processes are related through:

$$
dW^*_t = dW_t - X_tdt
$$

### Notes

-   $W_t$ has zero drift under $\bP$
-   $W^*_t$ has zero drift under $\bQ$
-   The definition $\bQ(A)$ is used so that for infinitesimal intervals,
    $d\bQ = \xi d \bP$
-   $\bQ(A) - E^P(1_A\xi_T) = \int 1_A \xi_T d\bP = \int_A \xi_T d\bP \rightarrow d\bQ = \xi d \bP$
-   TO remove drift, subracting the mean is usually not plausible since
    the risk premium may not be known. Sometimes switching to a
    risk-neutral setting is simpler. This is where the Girsanov Theorem
    can become a huge benefit.

# Examples <span style="color: red;">Consider moving this to another markdown file<span style="color: red;">

## Brownian Motion

$$
\Delta X_t \sim N(\mu\Delta, \sigma^2\Delta) \\
X_t \text{ is not a martingale because }E_t[X_t] = X_t \mu(T-t) \\
Z_t = X_t - \mu t \text{ is a martingale} \Rightarrow E_t[Z_{t+T} = Z_t]
$$

### Distribution of $W_5 - W_2$

$W_5 - W_2 \sim N(\mu = 0, \sigma^2 = 3)$

### Lognormal

Let $Y(t) = e^{cW(t)}$

Find $E[Y(t)]$

Note $cW(t) \sim N(c^2,t)$

Thus, $E_0[Y(t)] = \exp\left(\frac{c^2 t}{2}\right)$

## Forecasting

-   $\mu_t$ is the forecast of some random variable $Y_t$ for $t \leq T$
-   The information set grows over time so that
    $I_t \subseteq I_{t+1} \subseteq \dots \subseteq I_T$
-   $\mu_t = E^P[Y_T | I_t]$ is the forecast $\mu_t$ of the random
    variable $Y_T$ given the information set $I_t$
-   Let the sequence of forecasts $\mu_t$ w.r.t probability distribution
    P and information sets $I_t$ be a martingale. Then:

$$
E^P[\mu_{t+s} | I_t] = \mu_t \rightarrow \underbrace{E^P[E^P[Y_T | I_{t+s}]|I_t]}_{\text{Tower Property}} = E^P[Y_T | I_t] = \mu_t
$$

The eq. above states that the best forecast of a future forecast is what
we forecast now since that is all the info we have.

<span style="color: red;"><span style="color: red;">
